---
title: "Data Visualization for Time Series"
subtitle: "Fundação Getulio Vargas - EESP"
author: "Victor Henriques"
format: 
  revealjs:
    smaller: true
    toc: true
    toc-depth: 1
    theme: white
    slide-number: true
    incremental: false
    code-fold: false
    logo: logo.svg
knitr:
  opts_chunk:
    fig.width: 10
    fig.height: 6
    out.width: "100%"
    # out.height: "65vh"
    fig.align: "center"
editor: source
editor_options: 
  chunk_output_type: console
---

```{r}
library(yfR)
library(tidyverse)

tickers = c(
    "^BVSP"
    # , "ABEV3.SA", "ALOS3.SA", "ASAI3.SA", "B3SA3.SA",
    # "BBAS3.SA", "BBDC3.SA", "BBDC4.SA", "BBSE3.SA", "BEEF3.SA",
    # "BPAC11.SA", "BRAP4.SA", "BRAV3.SA", "BRFS3.SA", "BRKM5.SA",
    # "CCRO3.SA", "CMIG4.SA", "CMIN3.SA", "COGN3.SA", "CPFE3.SA",
    # "CPLE6.SA", "CRFB3.SA", "CSAN3.SA", "CSNA3.SA", "CVCB3.SA",
    # "CXSE3.SA", "CYRE3.SA", "ELET3.SA", "ELET6.SA", "EMBR3.SA",
    # "ENEV3.SA", "ENGI11.SA", "EQTL3.SA", "EZTC3.SA", "FLRY3.SA",
    # "GGBR4.SA", "GOAU4.SA", "HAPV3.SA", "HYPE3.SA", "IGTI11.SA",
    # "IRBR3.SA", "ISAE4.SA", "ITSA4.SA", "ITUB4.SA", "JBSS3.SA",
    # "KLBN11.SA", "LREN3.SA", "MGLU3.SA", "MRFG3.SA", "MRVE3.SA",
    # "MULT3.SA", "NTCO3.SA", "PCAR3.SA", "PETR3.SA",
    # "PETZ3.SA", "PRIO3.SA", "RADL3.SA", "RAIL3.SA", "RAIZ4.SA",
    # "RDOR3.SA", "RECV3.SA", "RENT3.SA", "RRRP3.SA", "SANB11.SA",
    # "SBSP3.SA", "SLCE3.SA", "SMTO3.SA", "SUZB3.SA", "TAEE11.SA",
    # "TIMS3.SA", "TOTS3.SA", "UGPA3.SA", "USIM5.SA", "VALE3.SA",
    # "VAMO3.SA", "VBBR3.SA", "WEGE3.SA", "YDUQ3.SA"
)

dados = yfR::yf_get(
  first_date = as.Date("2020-01-01"),
  last_date = Sys.Date(),
  tickers = tickers,
  bench_ticker = "^BVSP"
)
```

# Introduction

In today’s lecture, we will examine:

- **Outliers**: Where are the data located?
- **Dispersion**: How spread out are the data?
- **Shape**: What does the distribution look like?
- **Bivariate measures**: How do variables relate to each other?

Our goal is not only to calculate statistics, but to use visualization to reveal the stories contained in the data!

## Time Series Features

A time series is a sequence of data points recorded or measured in chronological order, usually at regular, equally spaced intervals.

- Trend: A trend exists when there is a long-term increase or decrease in the data. 

- Seasonality: Regular patterns that repeat over fixed periods like days, months, or years.

- Cyclic: Long-term ups and downs that do not follow a fixed seasonal schedule.

- Irregular: Random, unpredictable short-term changes

## Time plot

::: {.notes}

For time series data, the obvious graph to start with is a time plot. That is, the observations are plotted against the time of observation, with consecutive observations joined by straight lines.

:::


## Time Series - Seasonality 

```{r}
# ================================================================
# SEASONAL PLOT: AIRPASSENGERS
# Complete visualization for a Data Visualization class
# ================================================================

library(dplyr)
library(ggplot2)
library(scales)

# ================================================================
# 1. LOAD THE BUILT-IN AIRPASSENGERS DATA
# ================================================================

data("AirPassengers")

# ================================================================
# 2. CONVERT THE TIME SERIES INTO A DATA FRAME
#
# Each observation is described by:
# - calendar year;
# - seasonal position, represented by month;
# - number of passengers.
# ================================================================

air_seasonal_data <- tibble(
  Year = floor(
    as.numeric(
      time(AirPassengers)
    )
  ),

  Month_number = as.integer(
    cycle(AirPassengers)
  ),

  Passengers = as.numeric(
    AirPassengers
  )
) |>
  mutate(
    Month = factor(
      Month_number,

      levels = 1:12,

      labels = month.abb,

      ordered = TRUE
    ),

    Year_factor = factor(
      Year
    )
  )

# ================================================================
# 3. VALIDATE THE DATA STRUCTURE
# ================================================================

stopifnot(
  nrow(air_seasonal_data) ==
    length(AirPassengers),

  n_distinct(
    air_seasonal_data$Month
  ) == 12,

  all(
    air_seasonal_data |>
      count(Year) |>
      pull(n) == 12
  )
)

# ================================================================
# 4. CALCULATE TEACHING SUMMARY VALUES
# ================================================================

first_year <- min(
  air_seasonal_data$Year
)

last_year <- max(
  air_seasonal_data$Year
)

minimum_passengers <- min(
  air_seasonal_data$Passengers
)

maximum_passengers <- max(
  air_seasonal_data$Passengers
)

maximum_observation <- air_seasonal_data |>
  slice_max(
    order_by = Passengers,
    n = 1,
    with_ties = FALSE
  )

minimum_observation <- air_seasonal_data |>
  slice_min(
    order_by = Passengers,
    n = 1,
    with_ties = FALSE
  )

# ================================================================
# 5. CREATE DATA FOR THE FIRST AND LAST YEARS
#
# These lines are redrawn with greater emphasis to make the
# evolution of the seasonal pattern easier to recognize.
# ================================================================

first_year_data <- air_seasonal_data |>
  filter(
    Year == first_year
  )

last_year_data <- air_seasonal_data |>
  filter(
    Year == last_year
  )

# ================================================================
# 6. CREATE LABEL POSITIONS
# ================================================================

maximum_label <- maximum_observation |>
  mutate(
    Label = paste0(
      "Maximum: ",
      comma(Passengers),
      "\n",
      as.character(Month),
      " ",
      Year
    )
  )

minimum_label <- minimum_observation |>
  mutate(
    Label = paste0(
      "Minimum: ",
      comma(Passengers),
      "\n",
      as.character(Month),
      " ",
      Year
    )
  )

# ================================================================
# 7. BUILD THE SEASONAL PLOT
# ================================================================

air_seasonal_plot <- ggplot(
  data = air_seasonal_data,

  aes(
    x = Month,
    y = Passengers,
    group = Year,
    color = Year
  )
) +

  # --------------------------------------------------------------
  # Connect monthly observations within each year
  # --------------------------------------------------------------

  geom_line(
    linewidth = 0.85,
    alpha = 0.72,
    lineend = "round"
  ) +

  # --------------------------------------------------------------
  # Preserve the individual monthly observations
  # --------------------------------------------------------------

  geom_point(
    size = 1.9,
    alpha = 0.78
  ) +

  # --------------------------------------------------------------
  # Emphasize the first year
  # --------------------------------------------------------------

  geom_line(
    data = first_year_data,

    aes(
      x = Month,
      y = Passengers,
      group = Year
    ),

    inherit.aes = FALSE,

    color = "#64748B",
    linewidth = 1.15,
    linetype = "dashed",
    lineend = "round"
  ) +

  geom_point(
    data = first_year_data,

    aes(
      x = Month,
      y = Passengers
    ),

    inherit.aes = FALSE,

    color = "#64748B",
    fill = "white",
    shape = 21,
    size = 2.4,
    stroke = 0.8
  ) +

  # --------------------------------------------------------------
  # Emphasize the most recent year
  # --------------------------------------------------------------

  geom_line(
    data = last_year_data,

    aes(
      x = Month,
      y = Passengers,
      group = Year
    ),

    inherit.aes = FALSE,

    color = "#B2182B",
    linewidth = 1.35,
    lineend = "round"
  ) +

  geom_point(
    data = last_year_data,

    aes(
      x = Month,
      y = Passengers
    ),

    inherit.aes = FALSE,

    color = "#B2182B",
    fill = "white",
    shape = 21,
    size = 2.7,
    stroke = 0.9
  ) +

  # --------------------------------------------------------------
  # Highlight the maximum observed value
  # --------------------------------------------------------------

  geom_point(
    data = maximum_observation,

    aes(
      x = Month,
      y = Passengers
    ),

    inherit.aes = FALSE,

    color = "#B2182B",
    fill = "#FEE2E2",
    shape = 21,
    size = 4,
    stroke = 1.1
  ) +

  geom_label(
    data = maximum_label,

    aes(
      x = Month,
      y = Passengers,
      label = Label
    ),

    inherit.aes = FALSE,

    color = "#991B1B",
    fill = "white",
    label.size = 0.25,
    label.padding = grid::unit(
      0.18,
      "lines"
    ),
    size = 3,
    fontface = "bold",
    nudge_x = -0.55,
    nudge_y = 28
  ) +

  # --------------------------------------------------------------
  # Highlight the minimum observed value
  # --------------------------------------------------------------

  geom_point(
    data = minimum_observation,

    aes(
      x = Month,
      y = Passengers
    ),

    inherit.aes = FALSE,

    color = "#1D4ED8",
    fill = "#DBEAFE",
    shape = 21,
    size = 4,
    stroke = 1.1
  ) +

  geom_label(
    data = minimum_label,

    aes(
      x = Month,
      y = Passengers,
      label = Label
    ),

    inherit.aes = FALSE,

    color = "#1D4ED8",
    fill = "white",
    label.size = 0.25,
    label.padding = grid::unit(
      0.18,
      "lines"
    ),
    size = 3,
    fontface = "bold",
    nudge_x = 0.75,
    nudge_y = 24
  ) +

  # --------------------------------------------------------------
  # Continuous color scale represents chronological progression
  # --------------------------------------------------------------

  scale_color_gradientn(
    colors = c(
      "#CBD5E1",
      "#93C5FD",
      "#2F6FA5",
      "#173F73",
      "#B2182B"
    ),

    breaks = pretty(
      unique(
        air_seasonal_data$Year
      ),
      n = 6
    ),

    limits = c(
      first_year,
      last_year
    ),

    name = "Year"
  ) +

  # --------------------------------------------------------------
  # Vertical scale
  # --------------------------------------------------------------

  scale_y_continuous(
    labels = label_number(
      accuracy = 1,
      big.mark = ","
    ),

    expand = expansion(
      mult = c(
        0.04,
        0.14
      )
    )
  ) +

  # --------------------------------------------------------------
  # Titles and labels
  # --------------------------------------------------------------

  labs(
    title =
      "How Does Air Travel Vary Within Each Year?",

    subtitle = paste0(
      "Seasonal plot of monthly international airline passengers, ",
      first_year,
      "–",
      last_year
    ),

    x = "Month",

    y = "Passengers (thousands)",

    caption = paste0(
      "Each line represents one year. ",
      "The dashed gray line is ",
      first_year,
      "; the solid red line is ",
      last_year,
      "."
    )
  ) +

  # --------------------------------------------------------------
  # Presentation theme
  # --------------------------------------------------------------

  theme_minimal(
    base_size = 13
  ) +

  theme(
    plot.title = element_text(
      size = 20,
      face = "bold",
      color = "#172033",
      margin = margin(
        b = 4
      )
    ),

    plot.subtitle = element_text(
      size = 11.5,
      color = "#475569",
      margin = margin(
        b = 14
      )
    ),

    plot.caption = element_text(
      size = 8.5,
      color = "#64748B",
      hjust = 0,
      lineheight = 1.15,
      margin = margin(
        t = 10
      )
    ),

    axis.title = element_text(
      size = 11,
      face = "bold",
      color = "#334155"
    ),

    axis.text = element_text(
      size = 9,
      color = "#64748B"
    ),

    panel.grid.major.x = element_line(
      color = "#E5E7EB",
      linewidth = 0.4
    ),

    panel.grid.major.y = element_line(
      color = "#E5E7EB",
      linewidth = 0.4
    ),

    panel.grid.minor = element_blank(),

    legend.position = "right",

    legend.title = element_text(
      size = 10,
      face = "bold",
      color = "#334155"
    ),

    legend.text = element_text(
      size = 8.5,
      color = "#64748B"
    ),

    legend.key.height = grid::unit(
      1.2,
      "cm"
    ),

    plot.margin = margin(
      t = 10,
      r = 18,
      b = 8,
      l = 10
    )
  )

# ================================================================
# 8. DISPLAY THE FIGURE
# ================================================================

air_seasonal_plot
```

## Time Series - Seasonality 

```{r}
#| label: air-passengers-seasonal-subseries


# ================================================================
# SEASONAL SUBSERIES PLOT: AIRPASSENGERS
# Complete figure for a Data Visualization class
# ================================================================

library(dplyr)
library(ggplot2)
library(scales)

# ================================================================
# 1. LOAD THE BUILT-IN AIRPASSENGERS DATA
# ================================================================

data("AirPassengers")

# ================================================================
# 2. CONVERT THE TIME SERIES INTO A DATA FRAME
#
# AirPassengers is a monthly time series containing international
# airline passenger totals, measured in thousands.
# ================================================================

air_subseries_data <- tibble(
  Date = seq.Date(
    from = as.Date("1949-01-01"),
    by = "month",
    length.out = length(AirPassengers)
  ),

  Year = floor(
    as.numeric(
      time(AirPassengers)
    )
  ),

  Month_number = as.integer(
    cycle(AirPassengers)
  ),

  Passengers = as.numeric(
    AirPassengers
  )
) |>
  mutate(
    Month = factor(
      Month_number,
      levels = 1:12,
      labels = month.abb,
      ordered = TRUE
    )
  )

# ================================================================
# 3. VALIDATE THE TIME-SERIES STRUCTURE
# ================================================================

stopifnot(
  nrow(air_subseries_data) ==
    length(AirPassengers),

  n_distinct(
    air_subseries_data$Month
  ) == 12,

  all(
    air_subseries_data |>
      count(Year) |>
      pull(n) == 12
  )
)

# ================================================================
# 4. CALCULATE THE MEAN FOR EACH SEASONAL POSITION
#
# For monthly data, each seasonal position corresponds to a month.
# The horizontal reference line in each panel represents the mean
# passenger volume for that month across all years.
# ================================================================

monthly_means <- air_subseries_data |>
  group_by(Month) |>
  summarise(
    Monthly_mean = mean(
      Passengers,
      na.rm = TRUE
    ),

    Minimum = min(
      Passengers,
      na.rm = TRUE
    ),

    Maximum = max(
      Passengers,
      na.rm = TRUE
    ),

    Observations = n(),

    .groups = "drop"
  )

# ================================================================
# 5. JOIN THE MONTHLY MEANS TO THE OBSERVATIONS
# ================================================================

air_subseries_data <- air_subseries_data |>
  left_join(
    monthly_means |>
      select(
        Month,
        Monthly_mean
      ),
    by = "Month"
  )

# ================================================================
# 6. IDENTIFY THE FIRST AND LAST OBSERVATIONS IN EACH MONTH
#
# These points help demonstrate how each monthly subseries evolves
# from the beginning to the end of the sample.
# ================================================================

first_monthly_observations <- air_subseries_data |>
  group_by(Month) |>
  slice_min(
    order_by = Year,
    n = 1,
    with_ties = FALSE
  ) |>
  ungroup()

last_monthly_observations <- air_subseries_data |>
  group_by(Month) |>
  slice_max(
    order_by = Year,
    n = 1,
    with_ties = FALSE
  ) |>
  ungroup()

# ================================================================
# 7. CREATE LABELS FOR THE MONTHLY MEANS
# ================================================================

monthly_mean_labels <- monthly_means |>
  mutate(
    Mean_label = paste0(
      "Mean: ",
      comma(
        Monthly_mean,
        accuracy = 1
      )
    ),

    Label_year = max(
      air_subseries_data$Year
    ) + 0.35
  )

# ================================================================
# 8. CREATE THE SEASONAL SUBSERIES PLOT
# ================================================================

air_subseries_plot <- ggplot(
  data = air_subseries_data,
  aes(
    x = Year,
    y = Passengers
  )
) +

  # --------------------------------------------------------------
  # Monthly mean reference
  #
  # Each panel receives a different horizontal mean.
  # --------------------------------------------------------------

  geom_hline(
    data = monthly_means,
    aes(
      yintercept = Monthly_mean
    ),
    color = "#2F6FA5",
    linewidth = 0.9
  ) +

  # --------------------------------------------------------------
  # Time evolution within each seasonal position
  # --------------------------------------------------------------

  geom_line(
    color = "#64748B",
    linewidth = 0.75,
    alpha = 0.80,
    lineend = "round"
  ) +

  # --------------------------------------------------------------
  # Preserve the individual annual observations
  # --------------------------------------------------------------

  geom_point(
    color = "#64748B",
    fill = "white",
    shape = 21,
    size = 2.1,
    stroke = 0.65,
    alpha = 0.90
  ) +

  # --------------------------------------------------------------
  # Highlight the first observation in each month
  # --------------------------------------------------------------

  geom_point(
    data = first_monthly_observations,
    aes(
      x = Year,
      y = Passengers
    ),
    inherit.aes = FALSE,
    color = "#64748B",
    fill = "#F8FAFC",
    shape = 21,
    size = 2.7,
    stroke = 0.9
  ) +

  # --------------------------------------------------------------
  # Highlight the final observation in each month
  # --------------------------------------------------------------

  geom_point(
    data = last_monthly_observations,
    aes(
      x = Year,
      y = Passengers
    ),
    inherit.aes = FALSE,
    color = "#B2182B",
    fill = "#FEE2E2",
    shape = 21,
    size = 2.9,
    stroke = 1
  ) +

  # --------------------------------------------------------------
  # Label the monthly mean inside each panel
  # --------------------------------------------------------------

  geom_text(
    data = monthly_mean_labels,
    aes(
      x = Label_year,
      y = Monthly_mean,
      label = Mean_label
    ),
    inherit.aes = FALSE,
    color = "#1D4ED8",
    size = 2.7,
    fontface = "bold",
    hjust = 0,
    vjust = -0.35
  ) +

  # --------------------------------------------------------------
  # Create one mini time plot for each month
  # --------------------------------------------------------------

  facet_wrap(
    facets = vars(Month),
    ncol = 4,
    scales = "free_y",
    labeller = label_value
  ) +

  # --------------------------------------------------------------
  # Configure the year axis
  # --------------------------------------------------------------

  scale_x_continuous(
    breaks = c(
      1950,
      1955,
      1960
    ),

    limits = c(
      min(
        air_subseries_data$Year
      ),
      max(
        air_subseries_data$Year
      ) + 1.25
    ),

    expand = expansion(
      mult = c(
        0.02,
        0.01
      )
    )
  ) +

  # --------------------------------------------------------------
  # Configure the passenger axis
  # --------------------------------------------------------------

  scale_y_continuous(
    labels = label_number(
      accuracy = 1,
      big.mark = ","
    ),

    expand = expansion(
      mult = c(
        0.12,
        0.18
      )
    )
  ) +

  # --------------------------------------------------------------
  # Titles and labels
  # --------------------------------------------------------------

  labs(
    title =
      "How Does Each Month Change Over Time?",

    subtitle = paste0(
      "Seasonal subseries plot of AirPassengers | ",
      "Horizontal blue lines show the mean for each month"
    ),

    x = "Year",

    y = "Passengers (thousands)",

    caption = paste0(
      "Each panel contains observations from the same month across ",
      "different years. Gray lines show temporal evolution; ",
      "blue lines show monthly means; red points identify 1960."
    )
  ) +

  # --------------------------------------------------------------
  # Presentation theme
  # --------------------------------------------------------------

  theme_minimal(
    base_size = 12
  ) +

  theme(
    plot.title = element_text(
      size = 20,
      face = "bold",
      color = "#172033",
      margin = margin(
        b = 4
      )
    ),

    plot.subtitle = element_text(
      size = 11.5,
      color = "#475569",
      margin = margin(
        b = 14
      )
    ),

    plot.caption = element_text(
      size = 8.5,
      color = "#64748B",
      hjust = 0,
      lineheight = 1.15,
      margin = margin(
        t = 10
      )
    ),

    strip.background = element_rect(
      fill = "#F3F8FC",
      color = "#CBD5E1",
      linewidth = 0.55
    ),

    strip.text = element_text(
      size = 10.5,
      face = "bold",
      color = "#173F73",
      margin = margin(
        t = 5,
        r = 5,
        b = 5,
        l = 5
      )
    ),

    axis.title = element_text(
      size = 10.5,
      face = "bold",
      color = "#334155"
    ),

    axis.text = element_text(
      size = 8,
      color = "#64748B"
    ),

    panel.grid.major.x = element_line(
      color = "#E5E7EB",
      linewidth = 0.4
    ),

    panel.grid.major.y = element_line(
      color = "#E5E7EB",
      linewidth = 0.4
    ),

    panel.grid.minor = element_blank(),

    panel.spacing = grid::unit(
      0.8,
      "lines"
    ),

    plot.margin = margin(
      t = 10,
      r = 20,
      b = 8,
      l = 10
    )
  )

# ================================================================
# 9. DISPLAY THE FIGURE
# ================================================================

air_subseries_plot
```


## Box-Cox Transformation

**Box-Cox transformations** includes logarithmic and power
transformations controlled by $\lambda$

:::: {.columns}

::: {.column width="57%"}

### Definition

$$
\boxed{
w_t =
\begin{cases}
\log(y_t),
& \lambda = 0, \\[4pt]
\dfrac{
\operatorname{sign}(y_t)\lvert y_t\rvert^\lambda - 1
}{
\lambda
},
& \lambda \neq 0.
\end{cases}
}
$$

This modified formulation allows negative values of $y_t$, provided that

$$
\lambda > 0.
$$


:::

::: {.column width="43%"}

### Role of $\lambda$

$$
\begin{array}{ccl}
\lambda = 1
&\Rightarrow&
w_t = y_t - 1
\\[3pt]
\lambda = 0
&\Rightarrow&
w_t = \log(y_t)
\\[3pt]
\lambda \neq 0
&\Rightarrow&
\text{power transformation}
\end{array}
$$

- $\lambda = 1$ changes only the **level**

- $\lambda = 0$ applies a **log transformation**

- $\lambda \neq 0$ change the **shape and scale**

:::

::::

::: {.callout-tip appearance="simple" icon="false"}
Choose $\lambda$ to **stabilize the variance**, reduce scale-dependent
fluctuations, and make the variation more consistent over time.
:::

<small>
Box & Cox (1964); modified transformation from Bickel & Doksum (1981).
</small>

## Box-Cox Transformation

```{r}
# ============================================================
# BOX-COX TRANSFORMATION
# Illustration of variance stabilization over time
# ============================================================

# 1. Packages -------------------------------------------------

library(dplyr)
library(tidyr)
library(ggplot2)
library(slider)
library(forecast)
library(patchwork)
library(scales)

# 2. Reproducibility ------------------------------------------

set.seed(42)

# 3. Generate an illustrative positive time series ------------

n <- 180

data_boxcox <- tibble(
  Time = 1:n,

  # Exponentially increasing level
  Level = 18 * exp(0.012 * Time),

  # Multiplicative seasonal component
  Seasonality = 1 + 0.16 * sin(2 * pi * Time / 12),

  # Multiplicative random variation
  Noise = exp(rnorm(n, mean = 0, sd = 0.16)),

  # Observed positive series
  Original = Level * Seasonality * Noise
)

# 4. Estimate the Box-Cox lambda -------------------------------

lambda <- forecast::BoxCox.lambda(
  x      = ts(data_boxcox$Original, frequency = 12),
  method = "guerrero",
  lower  = -2,
  upper  = 2
)

# 5. Apply the Box-Cox transformation --------------------------

data_boxcox <- data_boxcox |>
  mutate(
    BoxCox = forecast::BoxCox(
      Original,
      lambda = lambda
    )
  )

# 6. Calculate rolling standard deviations --------------------

rolling_window <- 18

data_boxcox <- data_boxcox |>
  mutate(
    SD_Original = slide_dbl(
      Original,
      sd,
      .before   = floor(rolling_window / 2),
      .after    = floor(rolling_window / 2) - 1,
      .complete = TRUE
    ),

    SD_BoxCox = slide_dbl(
      BoxCox,
      sd,
      .before   = floor(rolling_window / 2),
      .after    = floor(rolling_window / 2) - 1,
      .complete = TRUE
    ),

    Original_Lower = Original - SD_Original,
    Original_Upper = Original + SD_Original,

    BoxCox_Lower = BoxCox - SD_BoxCox,
    BoxCox_Upper = BoxCox + SD_BoxCox
  )

# 7. Compare early and late variability ------------------------

comparison_window <- 45

sd_early_original <- data_boxcox |>
  slice_head(n = comparison_window) |>
  summarise(value = sd(Original)) |>
  pull(value)

sd_late_original <- data_boxcox |>
  slice_tail(n = comparison_window) |>
  summarise(value = sd(Original)) |>
  pull(value)

sd_early_boxcox <- data_boxcox |>
  slice_head(n = comparison_window) |>
  summarise(value = sd(BoxCox)) |>
  pull(value)

sd_late_boxcox <- data_boxcox |>
  slice_tail(n = comparison_window) |>
  summarise(value = sd(BoxCox)) |>
  pull(value)

ratio_original <- sd_late_original / sd_early_original
ratio_boxcox   <- sd_late_boxcox / sd_early_boxcox

# 8. Color palette ---------------------------------------------

color_navy  <- "#17324D"
color_teal  <- "#00A6A6"
color_coral <- "#EF6C5B"
color_grey  <- "#6B7280"
color_panel <- "#F4F7FA"

# 9. Common theme ----------------------------------------------

theme_boxcox <- theme_minimal(base_size = 13) +
  theme(
    plot.background = element_rect(
      fill  = "white",
      color = NA
    ),
    panel.background = element_rect(
      fill  = color_panel,
      color = NA
    ),
    panel.grid.major.x = element_blank(),
    panel.grid.minor   = element_blank(),
    panel.grid.major.y = element_line(
      color     = "white",
      linewidth = 0.8
    ),
    axis.title = element_text(
      color = color_navy,
      face  = "bold"
    ),
    axis.text = element_text(
      color = color_grey
    ),
    axis.ticks = element_blank(),
    plot.title = element_text(
      color  = color_navy,
      face   = "bold",
      size   = 16,
      margin = margin(b = 8)
    ),
    plot.subtitle = element_text(
      color  = color_grey,
      size   = 11,
      margin = margin(b = 10)
    ),
    plot.caption = element_text(
      color  = color_grey,
      size   = 9,
      hjust  = 0
    ),
    plot.margin = margin(
      t = 15,
      r = 20,
      b = 10,
      l = 15
    )
  )

# 10. Original-scale plot --------------------------------------

plot_original <- ggplot(
  data_boxcox,
  aes(x = Time, y = Original)
) +
  annotate(
    "rect",
    xmin  = 1,
    xmax  = comparison_window,
    ymin  = -Inf,
    ymax  = Inf,
    fill  = color_navy,
    alpha = 0.035
  ) +
  annotate(
    "rect",
    xmin  = n - comparison_window + 1,
    xmax  = n,
    ymin  = -Inf,
    ymax  = Inf,
    fill  = color_navy,
    alpha = 0.035
  ) +
  geom_ribbon(
    aes(
      ymin = Original_Lower,
      ymax = Original_Upper
    ),
    fill  = color_coral,
    alpha = 0.14,
    na.rm = TRUE
  ) +
  geom_line(
    color     = color_coral,
    linewidth = 1
  ) +
  annotate(
    "label",
    x = Inf,
    y = Inf,
    label = paste0(
      "Late / early SD = ",
      number(ratio_original, accuracy = 0.1),
      "\u00d7"
    ),
    hjust = 1.08,
    vjust = 1.4,
    color = color_coral,
    fill  = "white",
    fontface = "bold",
    label.size = 0,
    size = 4
  ) +
  labs(
    title = "Before: fluctuations grow with the level",
    subtitle = paste0(
      "The magnitude of the variation is dependent ",
      "on the scale of the series"
    ),
    x = NULL,
    y = "Original scale"
  ) +
  scale_x_continuous(
    breaks = seq(0, n, by = 30),
    expand = expansion(mult = c(0.01, 0.01))
  ) +
  scale_y_continuous(
    labels = label_number(accuracy = 1),
    expand = expansion(mult = c(0.08, 0.12))
  ) +
  theme_boxcox +
  theme(
    axis.text.x  = element_blank(),
    axis.title.x = element_blank()
  )

# 11. Box-Cox-transformed plot ---------------------------------

plot_transformed <- ggplot(
  data_boxcox,
  aes(x = Time, y = BoxCox)
) +
  annotate(
    "rect",
    xmin  = 1,
    xmax  = comparison_window,
    ymin  = -Inf,
    ymax  = Inf,
    fill  = color_navy,
    alpha = 0.035
  ) +
  annotate(
    "rect",
    xmin  = n - comparison_window + 1,
    xmax  = n,
    ymin  = -Inf,
    ymax  = Inf,
    fill  = color_navy,
    alpha = 0.035
  ) +
  geom_ribbon(
    aes(
      ymin = BoxCox_Lower,
      ymax = BoxCox_Upper
    ),
    fill  = color_teal,
    alpha = 0.14,
    na.rm = TRUE
  ) +
  geom_line(
    color     = color_teal,
    linewidth = 1
  ) +
  annotate(
    "label",
    x = Inf,
    y = Inf,
    label = paste0(
      "Late / early SD = ",
      number(ratio_boxcox, accuracy = 0.1),
      "\u00d7"
    ),
    hjust = 1.08,
    vjust = 1.4,
    color = color_teal,
    fill  = "white",
    fontface = "bold",
    label.size = 0,
    size = 4
  ) +
  labs(
    title = paste0(
      "After Box-Cox: variation is more consistent over time",
      "  (\u03bb = ",
      number(lambda, accuracy = 0.01),
      ")"
    ),
    subtitle = paste0(
      "Large values are compressed more strongly, ",
      "reducing scale-dependent fluctuations"
    ),
    x = "Time",
    y = "Transformed scale"
  ) +
  scale_x_continuous(
    breaks = seq(0, n, by = 30),
    expand = expansion(mult = c(0.01, 0.01))
  ) +
  scale_y_continuous(
    labels = label_number(accuracy = 0.01),
    expand = expansion(mult = c(0.08, 0.12))
  ) +
  theme_boxcox

# 12. Combine plots --------------------------------------------

final_plot <- plot_original / plot_transformed +
  plot_annotation(
    title = paste0(
      "Box-Cox transformation stabilizes ",
      "scale-dependent variance"
    ),
    subtitle = paste0(
      "Shaded ribbons represent an ",
      rolling_window,
      "-period rolling standard deviation"
    ),
    caption = paste0(
      "Illustrative synthetic series. ",
      "Box-Cox requires strictly positive observations."
    ),
    theme = theme(
      plot.title = element_text(
        color  = color_navy,
        face   = "bold",
        size   = 21
      ),
      plot.subtitle = element_text(
        color  = color_grey,
        size   = 12
      ),
      plot.caption = element_text(
        color = color_grey,
        size  = 9,
        hjust = 0
      ),
      plot.margin = margin(
        t = 15,
        r = 20,
        b = 15,
        l = 15
      )
    )
  )

# 13. Display --------------------------------------------------

final_plot
```


## Index

```{r}
# ================================================================
# DAILY TO MONTHLY DATA
# Monthly prices and compounded monthly returns
# ================================================================

library(dplyr)
library(lubridate)

# ================================================================
# 1. IDENTIFY THE PRICE COLUMN
#
# The code searches for common price-column names.
# Place the exact name used in your dataset first if necessary.
# ================================================================

price_candidates <- c(
  "adjusted_prices",
  "adjusted_price",
  "price_adjusted",
  "adj_price",
  "adjusted_close",
  "price",
  "Price",
  "close",
  "Close"
)

price_column <- intersect(
  price_candidates,
  names(dados)
)

if (length(price_column) == 0L) {
  stop(
    paste0(
      "No price column was found. Available columns are: ",
      paste(
        names(dados),
        collapse = ", "
      )
    )
  )
}

price_column <- price_column[1]

# ================================================================
# 2. VALIDATE THE REQUIRED COLUMNS
# ================================================================

required_columns <- c(
  "ticker",
  "ref_date",
  "ret_adjusted_prices"
)

missing_columns <- setdiff(
  required_columns,
  names(dados)
)

if (length(missing_columns) > 0L) {
  stop(
    paste0(
      "The following required columns are missing: ",
      paste(
        missing_columns,
        collapse = ", "
      )
    )
  )
}

# ================================================================
# 3. HELPER FUNCTION
#
# Returns the last finite observation in a vector.
# ================================================================

last_finite <- function(x) {

  x <- as.numeric(x)

  valid_positions <- which(
    is.finite(x)
  )

  if (length(valid_positions) == 0L) {
    return(NA_real_)
  }

  x[
    valid_positions[
      length(valid_positions)
    ]
  ]
}

# ================================================================
# 4. PREPARE THE DAILY DATA
#
# The price column is renamed to adjusted_price.
# The return column is renamed to adjusted_return.
# ================================================================

daily_data <- dados |>
  transmute(
    ticker = as.character(ticker),
    ref_date = as.Date(ref_date),
    adjusted_price = suppressWarnings(
      as.numeric(
        .data[[price_column]]
      )
    ),

    adjusted_return = suppressWarnings(
      as.numeric(
        ret_adjusted_prices
      )
    )
  ) |>
  filter(
    !is.na(ticker),
    ticker != "",
    !is.na(ref_date)
  ) |>
  arrange(
    ticker,
    ref_date
  )

if (nrow(daily_data) == 0L) {
  stop(
    "No valid ticker and date observations were found."
  )
}

# ================================================================
# 5. AGGREGATE DUPLICATE TICKER-DATE OBSERVATIONS
#
# If duplicate observations exist on the same date, the mean of
# the finite values is used.
# ================================================================

daily_data <- daily_data |>
  group_by(
    ticker,
    ref_date
  ) |>
  summarise(
    adjusted_price = {
      valid_values <- adjusted_price[
        is.finite(adjusted_price)
      ]
      if (length(valid_values) == 0L) {
        NA_real_
      } else {
        mean(valid_values)
      }
    },
    adjusted_return = {
      valid_values <- adjusted_return[
        is.finite(adjusted_return)
      ]
      if (length(valid_values) == 0L) {
        NA_real_
      } else {
        mean(valid_values)
      }
    },

    .groups = "drop"
  ) |>
  mutate(
    month = floor_date(
      ref_date,
      unit = "month"
    )
  ) |>
  arrange(
    ticker,
    ref_date
  )

# ================================================================
# 6. CREATE MONTHLY PRICES
#
# The monthly price is the last valid price observed in the month.
# The actual observation date is preserved.
# ================================================================

monthly_prices <- daily_data |>
  filter(
    is.finite(adjusted_price)
  ) |>
  group_by(
    ticker,
    month
  ) |>
  arrange(
    ref_date,
    .by_group = TRUE
  ) |>
  summarise(
    monthly_price_date = ref_date[
      length(ref_date)
    ],

    monthly_adjusted_price = last_finite(
      adjusted_price
    ),

    first_adjusted_price = adjusted_price[1],

    average_adjusted_price = mean(
      adjusted_price
    ),

    minimum_adjusted_price = min(
      adjusted_price
    ),

    maximum_adjusted_price = max(
      adjusted_price
    ),

    price_observations = n(),

    .groups = "drop"
  ) |>
  arrange(
    ticker,
    month
  )

if (nrow(monthly_prices) == 0L) {
  stop(
    paste0(
      "No finite prices were found in column '",
      price_column,
      "'."
    )
  )
}

# ================================================================
# 7. CALCULATE MONTHLY RETURNS FROM MONTH-END PRICES
#
# R_m = P_m / P_(m-1) - 1
# ================================================================

monthly_prices <- monthly_prices |>
  group_by(ticker) |>
  arrange(
    month,
    .by_group = TRUE
  ) |>
  mutate(
    previous_monthly_price = lag(
      monthly_adjusted_price
    ),

    monthly_return_from_price = if_else(
      is.finite(previous_monthly_price) &
        previous_monthly_price != 0,

      monthly_adjusted_price /
        previous_monthly_price -
        1,

      NA_real_
    ),

    monthly_log_return_from_price = if_else(
      is.finite(previous_monthly_price) &
        previous_monthly_price > 0 &
        monthly_adjusted_price > 0,

      log(
        monthly_adjusted_price /
          previous_monthly_price
      ),

      NA_real_
    )
  ) |>
  ungroup()

# ================================================================
# 8. COMPOUND DAILY SIMPLE RETURNS WITHIN EACH MONTH
#
# R_m = product(1 + r_d) - 1
#
# This calculation assumes that ret_adjusted_prices contains
# decimal simple returns:
#
#  0.01  =  1%
# -0.02  = -2%
# ================================================================

monthly_returns <- daily_data |>
  group_by(
    ticker,
    month
  ) |>
  summarise(
    valid_daily_returns = sum(
      is.finite(adjusted_return)
    ),

    first_return_date = {
      valid_dates <- ref_date[
        is.finite(adjusted_return)
      ]

      if (length(valid_dates) == 0L) {
        as.Date(NA)
      } else {
        min(valid_dates)
      }
    },

    last_return_date = {
      valid_dates <- ref_date[
        is.finite(adjusted_return)
      ]

      if (length(valid_dates) == 0L) {
        as.Date(NA)
      } else {
        max(valid_dates)
      }
    },

    monthly_adjusted_return = {
      valid_returns <- adjusted_return[
        is.finite(adjusted_return)
      ]

      if (length(valid_returns) == 0L) {
        NA_real_
      } else if (any(valid_returns <= -1)) {
        NA_real_
      } else {
        prod(
          1 + valid_returns
        ) - 1
      }
    },

    monthly_log_return = {
      valid_returns <- adjusted_return[
        is.finite(adjusted_return) &
          adjusted_return > -1
      ]

      if (length(valid_returns) == 0L) {
        NA_real_
      } else {
        sum(
          log1p(valid_returns)
        )
      }
    },

    .groups = "drop"
  ) |>
  arrange(
    ticker,
    month
  ) |>
  mutate(
    monthly_simple_return_from_logs = if_else(
      is.finite(monthly_log_return),

      expm1(
        monthly_log_return
      ),

      NA_real_
    )
  )

# ================================================================
# 9. COMBINE MONTHLY PRICES AND RETURNS
# ================================================================

monthly_data <- monthly_prices |>
  full_join(
    monthly_returns,
    by = c(
      "ticker",
      "month"
    )
  ) |>
  arrange(
    ticker,
    month
  )

# ================================================================
# 10. COMPARE THE TWO MONTHLY-RETURN CALCULATIONS
#
# The two returns should be close when:
#
# - daily returns were calculated from the same adjusted prices;
# - the first daily return captures the movement from the final
#   price of the previous month;
# - no relevant daily observations are missing;
# - both series use the same adjustment methodology.
# ================================================================

monthly_data <- monthly_data |>
  mutate(
    return_difference =
      monthly_adjusted_return -
      monthly_return_from_price,

    absolute_return_difference = abs(
      return_difference
    ),

    return_calculations_match = case_when(
      is.na(monthly_adjusted_return) |
        is.na(monthly_return_from_price) ~
        NA,

      absolute_return_difference <
        1e-8 ~
        TRUE,

      TRUE ~
        FALSE
    )
  )

# ================================================================
# 11. VALIDATE THE MONTHLY DATASET
# ================================================================

duplicated_months <- monthly_data |>
  count(
    ticker,
    month
  ) |>
  filter(
    n > 1L
  )

if (nrow(duplicated_months) > 0L) {
  stop(
    "The monthly dataset contains duplicated ticker-month pairs."
  )
}

stopifnot(
  all(
    monthly_data$monthly_adjusted_price[
      !is.na(
        monthly_data$monthly_adjusted_price
      )
    ] > 0
  )
)

# ================================================================
# 12. FINAL MONTHLY DATASET
# ================================================================

monthly_data <- monthly_data |>
  select(
    ticker,
    month,

    monthly_price_date,
    monthly_adjusted_price,

    first_adjusted_price,
    average_adjusted_price,
    minimum_adjusted_price,
    maximum_adjusted_price,
    price_observations,

    previous_monthly_price,
    monthly_return_from_price,
    monthly_log_return_from_price,

    first_return_date,
    last_return_date,
    valid_daily_returns,
    monthly_adjusted_return,
    monthly_log_return,
    monthly_simple_return_from_logs,

    return_difference,
    absolute_return_difference,
    return_calculations_match
  )

# ================================================================
# 13. DISPLAY THE COMPLETE RESULT
# ================================================================

monthly_data
```



## Technical Indicators

The simple moving average assigns equal weight to the $w$ most recent observations:

$$
\operatorname{MA}_{t}^{(w)}
=
\frac{1}{w}
\sum_{j=0}^{w-1}Y_{t-j}.
$$

<!-- Each observation remains in the calculation for exactly $w$ periods and receives weight $1/w$ -->

The exponential moving average assigns greater weight to more recent observations:

$$
\operatorname{EMA}_{t}
=
\alpha Y_t
+
(1-\alpha)\operatorname{EMA}_{t-1},
\qquad
0<\alpha\leq1.
$$

<!-- A larger $\alpha$ responds more quickly to recent changes, while a smaller $\alpha$ produces a smoother estimate -->

> **The same weighting principles can be applied to other local measures. Moving means track changes in local position, while moving standard deviations track changes in local dispersion.**


## Technical Indicators

```{r}
library(dplyr)
library(ggplot2)
library(scales)

# ================================================================
# 1. PARAMETERS
# ================================================================

window_size <- 21L

# Conventional equivalence between an EMA span and alpha
alpha <- 2 / (window_size + 1)

# ================================================================
# 2. IDENTIFY THE PRICE COLUMN
#
# The first matching column is selected.
# Add the actual name to this vector if necessary.
# ================================================================

price_candidates <- c(
  "adjusted_prices",
  "adjusted_price",
  "price_adjusted",
  "adj_price",
  "adjusted_close",
  "price",
  "Price",
  "close",
  "Close"
)

price_column <- price_candidates[
  price_candidates %in% names(dados)
][1]

if (
  length(price_column) == 0 ||
  is.na(price_column)
) {
  stop(
    paste0(
      "No price column was found. Available columns are: ",
      paste(
        names(dados),
        collapse = ", "
      )
    )
  )
}

# ================================================================
# 3. SELECT THE FIRST VALID TICKER
# ================================================================

ticker_selected <- dados |>
  filter(
    !is.na(ticker)
  ) |>
  distinct(ticker) |>
  slice_head(n = 1) |>
  pull(ticker) |>
  as.character()

if (
  length(ticker_selected) == 0 ||
  is.na(ticker_selected)
) {
  stop(
    "No valid ticker was found in the dataset."
  )
}

# ================================================================
# 4. PREPARE THE PRICE SERIES
#
# If duplicate prices exist for the same date and ticker,
# their mean is used to create one price per date.
# ================================================================

technical_data <- dados |>
  filter(
    ticker == ticker_selected,
    !is.na(ref_date),
    !is.na(.data[[price_column]]),
    is.finite(
      as.numeric(
        .data[[price_column]]
      )
    )
  ) |>
  transmute(
    ticker = as.character(ticker),
    ref_date = as.Date(ref_date),
    price = as.numeric(
      .data[[price_column]]
    )
  ) |>
  group_by(
    ticker,
    ref_date
  ) |>
  summarise(
    price = mean(
      price,
      na.rm = TRUE
    ),
    .groups = "drop"
  ) |>
  arrange(ref_date) %>% 
  filter(ref_date > as.Date("2025-01-01"))

n_observations <- nrow(
  technical_data
)

if (n_observations < window_size) {
  stop(
    paste0(
      "At least ",
      window_size,
      " valid price observations are required for ",
      ticker_selected,
      ". Only ",
      n_observations,
      " were found."
    )
  )
}

# ================================================================
# 5. TRAILING SIMPLE MOVING AVERAGE
#
# MA_t = mean of the latest w observations, including time t.
# ================================================================

trailing_mean <- function(x, window) {

  n <- length(x)

  result <- rep(
    NA_real_,
    n
  )

  if (n >= window) {

    for (i in seq.int(window, n)) {

      result[i] <- mean(
        x[(i - window + 1):i],
        na.rm = TRUE
      )
    }
  }

  result
}

# ================================================================
# 6. TRAILING ROLLING STANDARD DEVIATION
#
# The sample standard deviation is calculated over the same
# trailing window used by the simple moving average.
# ================================================================

trailing_sd <- function(x, window) {

  n <- length(x)

  result <- rep(
    NA_real_,
    n
  )

  if (n >= window) {

    for (i in seq.int(window, n)) {

      result[i] <- sd(
        x[(i - window + 1):i],
        na.rm = TRUE
      )
    }
  }

  result
}

# ================================================================
# 7. EXPONENTIAL MOVING AVERAGE
#
# EMA_t = alpha * P_t + (1 - alpha) * EMA_(t-1)
# ================================================================

exponential_mean <- function(x, alpha) {

  n <- length(x)

  result <- rep(
    NA_real_,
    n
  )

  result[1] <- x[1]

  if (n >= 2) {

    for (i in 2:n) {

      result[i] <-
        alpha * x[i] +
        (1 - alpha) * result[i - 1]
    }
  }

  result
}

# ================================================================
# 8. EXPONENTIALLY WEIGHTED STANDARD DEVIATION
#
# The recursive exponential variance is:
#
# V_t = (1 - alpha) *
#       [V_(t-1) + alpha * (P_t - EMA_(t-1))^2]
#
# EWSD_t = sqrt(V_t)
# ================================================================

exponential_sd <- function(x, alpha) {

  n <- length(x)

  exponential_average <- rep(
    NA_real_,
    n
  )

  exponential_variance <- rep(
    NA_real_,
    n
  )

  exponential_average[1] <- x[1]

  exponential_variance[1] <- 0

  if (n >= 2) {

    for (i in 2:n) {

      previous_mean <-
        exponential_average[i - 1]

      previous_variance <-
        exponential_variance[i - 1]

      exponential_average[i] <-
        alpha * x[i] +
        (1 - alpha) * previous_mean

      exponential_variance[i] <-
        (1 - alpha) * (
          previous_variance +
            alpha *
              (x[i] - previous_mean)^2
        )
    }
  }

  sqrt(
    pmax(
      exponential_variance,
      0
    )
  )
}

# ================================================================
# 9. CALCULATE ALL PRICE INDICATORS
# ================================================================

technical_data <- technical_data |>
  mutate(
    MA = trailing_mean(
      price,
      window = window_size
    ),

    EMA = exponential_mean(
      price,
      alpha = alpha
    ),

    Rolling_SD = trailing_sd(
      price,
      window = window_size
    ),

    Exponential_SD = exponential_sd(
      price,
      alpha = alpha
    ),

    MA_lower = MA - Rolling_SD,

    MA_upper = MA + Rolling_SD,

    EMA_lower = EMA - Exponential_SD,

    EMA_upper = EMA + Exponential_SD
  )

# ================================================================
# 10. VALIDATE THE INDICATORS
# ================================================================

expected_complete_windows <-
  n_observations -
  window_size +
  1L

stopifnot(
  nrow(technical_data) ==
    n_observations,

  sum(
    !is.na(
      technical_data$MA
    )
  ) ==
    expected_complete_windows,

  sum(
    !is.na(
      technical_data$Rolling_SD
    )
  ) ==
    expected_complete_windows,

  all(
    technical_data$Rolling_SD[
      !is.na(
        technical_data$Rolling_SD
      )
    ] >= 0
  ),

  all(
    technical_data$Exponential_SD[
      !is.na(
        technical_data$Exponential_SD
      )
    ] >= 0
  )
)

# ================================================================
# 11. DETERMINE A SUITABLE PRICE FORMAT
#
# Large price values are displayed with no decimals.
# Smaller values retain greater precision.
# ================================================================

median_price <- median(
  technical_data$price,
  na.rm = TRUE
)

price_accuracy <- case_when(
  median_price >= 100 ~ 1,
  median_price >= 10 ~ 0.1,
  TRUE ~ 0.01
)

price_label_function <- label_number(
  accuracy = price_accuracy,
  big.mark = ","
)

# ================================================================
# 12. CREATE THE SINGLE FIGURE
# ================================================================

technical_indicators_plot <- ggplot(
  data = technical_data,
  aes(x = ref_date)
) +

  # --------------------------------------------------------------
  # SIMPLE MOVING-AVERAGE DISPERSION BAND
  #
  # MA_t plus or minus rolling SD_t
  # --------------------------------------------------------------

  geom_ribbon(
    aes(
      ymin = MA_lower,
      ymax = MA_upper,
      fill = "MA ± rolling SD"
    ),
    alpha = 0.17,
    na.rm = TRUE
  ) +

  # --------------------------------------------------------------
  # EXPONENTIALLY WEIGHTED DISPERSION BAND
  #
  # EMA_t plus or minus exponentially weighted SD_t
  # --------------------------------------------------------------

  geom_ribbon(
    aes(
      ymin = EMA_lower,
      ymax = EMA_upper,
      fill = "EMA ± exponentially weighted SD"
    ),
    alpha = 0.13,
    na.rm = TRUE
  ) +

  # --------------------------------------------------------------
  # OBSERVED PRICE
  # --------------------------------------------------------------

  geom_line(
    aes(
      y = price,
      color = "Observed price",
      linewidth = "Observed price"
    ),
    alpha = 0.58,
    lineend = "round",
    na.rm = TRUE
  ) +

  # --------------------------------------------------------------
  # SIMPLE MOVING AVERAGE
  # --------------------------------------------------------------

  geom_line(
    aes(
      y = MA,
      color = "Simple moving average",
      linewidth = "Simple moving average"
    ),
    lineend = "round",
    na.rm = TRUE
  ) +

  # --------------------------------------------------------------
  # EXPONENTIAL MOVING AVERAGE
  # --------------------------------------------------------------

  geom_line(
    aes(
      y = EMA,
      color = "Exponential moving average",
      linewidth = "Exponential moving average"
    ),
    lineend = "round",
    na.rm = TRUE
  ) +

  # --------------------------------------------------------------
  # LINE COLORS
  # --------------------------------------------------------------

  scale_color_manual(
    name = NULL,

    values = c(
      "Observed price" = "#64748B",
      "Simple moving average" = "#2F6FA5",
      "Exponential moving average" = "#D97706"
    ),

    breaks = c(
      "Observed price",
      "Simple moving average",
      "Exponential moving average"
    )
  ) +

  # --------------------------------------------------------------
  # BAND COLORS
  # --------------------------------------------------------------

  scale_fill_manual(
    name = NULL,

    values = c(
      "MA ± rolling SD" = "#2F6FA5",
      "EMA ± exponentially weighted SD" = "#D97706"
    ),

    breaks = c(
      "MA ± rolling SD",
      "EMA ± exponentially weighted SD"
    )
  ) +

  # --------------------------------------------------------------
  # LINE WIDTHS
  # --------------------------------------------------------------

  scale_linewidth_manual(
    name = NULL,

    values = c(
      "Observed price" = 0.55,
      "Simple moving average" = 1.10,
      "Exponential moving average" = 1.10
    )
  ) +

  # --------------------------------------------------------------
  # DATE AXIS
  # --------------------------------------------------------------

  scale_x_date(
    date_breaks = "6 months",

    date_minor_breaks = "3 months",

    date_labels = "%b\n%Y",

    expand = expansion(
      mult = c(
        0.01,
        0.02
      )
    )
  ) +

  # --------------------------------------------------------------
  # PRICE AXIS
  # --------------------------------------------------------------

  scale_y_continuous(
    labels = price_label_function,

    expand = expansion(
      mult = c(
        0.06,
        0.10
      )
    )
  ) +

  # --------------------------------------------------------------
  # TITLES
  # --------------------------------------------------------------

  labs(
    title =
      "How Do Moving Indicators Track Local Price and Dispersion?",

    subtitle = paste0(
      ticker_selected,
      " price | ",
      window_size,
      "-observation trailing window | ",
      "EMA alpha = ",
      number(
        alpha,
        accuracy = 0.001
      )
    ),

    x = "Date",

    y = "Price",

    caption = paste0(
      "Solid lines represent observed and smoothed prices. ",
      "Shaded areas represent one local standard deviation ",
      "above and below each moving average."
    )
  ) +

  # --------------------------------------------------------------
  # PRESENTATION THEME
  # --------------------------------------------------------------

  theme_minimal(
    base_size = 13
  ) +

  theme(
    plot.title = element_text(
      size = 20,
      face = "bold",
      color = "#172033",
      margin = margin(
        b = 4
      )
    ),

    plot.subtitle = element_text(
      size = 11.5,
      color = "#475569",
      margin = margin(
        b = 13
      )
    ),

    plot.caption = element_text(
      size = 8.5,
      color = "#64748B",
      hjust = 0,
      lineheight = 1.15,
      margin = margin(
        t = 10
      )
    ),

    axis.title = element_text(
      size = 11,
      face = "bold",
      color = "#334155"
    ),

    axis.text = element_text(
      size = 8.5,
      color = "#64748B"
    ),

    panel.grid.major = element_line(
      color = "#E5E7EB",
      linewidth = 0.4
    ),

    panel.grid.minor.x = element_line(
      color = "#F1F5F9",
      linewidth = 0.25
    ),

    panel.grid.minor.y = element_blank(),

    legend.position = "bottom",

    legend.box = "vertical",

    legend.text = element_text(
      size = 9,
      color = "#475569"
    ),

    legend.key.width = grid::unit(
      1.6,
      "cm"
    ),

    plot.margin = margin(
      t = 10,
      r = 20,
      b = 8,
      l = 10
    )
  ) +

  guides(
    color = guide_legend(
      order = 1,
      nrow = 1,
      override.aes = list(
        alpha = 1
      )
    ),

    fill = guide_legend(
      order = 2,
      nrow = 1
    ),

    linewidth = "none"
  )

# ================================================================
# 13. DISPLAY THE FIGURE
# ================================================================

technical_indicators_plot
```

## Time Series Decomposition

A time series $Y_t$ might be decomposed into three additive components

$$
Y_t = T_t +S_t + R_t
$$
where $T_t$, $S_t$, and $R_t$ are the trend, seasonal and the remainder, respectively.

Let $m$ denote the seasonal period. The components can be estimated as follows:

1. Trend: $\hat{T}_t = \frac{1}{m} \sum_{j=-\lfloor m/2 \rfloor}^{\lfloor m/2 \rfloor} Y_{t+j}$, where a centered moving average smooths short-term and seasonal fluctuations.

2. Detrended series: $D_t = Y_t-\widehat{T}_t.$ For each seasonal position $k=1,\ldots,m$.

3. Seasonal component:$\widetilde{S}_k = \frac{1}{n_k} \sum_{t:\,s(t)=k} D_t$, and centering the resulting seasonal indices: $\widehat{S}_k = \widetilde{S}_k - \frac{1}{m} \sum_{\ell=1}^{m}\widetilde{S}_{\ell}, \qquad \sum_{k=1}^{m}\widehat{S}_k=0.$

4. Remainder: $\widehat{R}_t = Y_t-\widehat{T}_t-\widehat{S}_{s(t)}$  

## Time Series Decomposition

```{r}
library(ggplot2)
library(dplyr)
library(tidyr)
library(scales)

# Load the built-in monthly time series
data("AirPassengers")

# Convert the time-series index to monthly dates
air_data <- tibble(
  Date = seq.Date(
    from = as.Date("1949-01-01"),
    by = "month",
    length.out = length(AirPassengers)
  ),
  Passengers = as.numeric(AirPassengers),
  Log_Passengers = log(as.numeric(AirPassengers))
)

# STL decomposition on the log scale
# The log transformation stabilizes the increasing seasonal amplitude
air_stl <- stl(
  ts(
    air_data$Log_Passengers,
    start = c(1949, 1),
    frequency = 12
  ),
  s.window = "periodic",
  robust = TRUE
)

# Extract the STL components
air_components <- air_data |>
  mutate(
    Observed = Log_Passengers,
    Trend = as.numeric(
      air_stl$time.series[, "trend"]
    ),
    Seasonal = as.numeric(
      air_stl$time.series[, "seasonal"]
    ),
    Remainder = as.numeric(
      air_stl$time.series[, "remainder"]
    ),
    Reconstructed = Trend + Seasonal + Remainder
  )

# Validate the additive identity
reconstruction_error <- max(
  abs(
    air_components$Observed -
      air_components$Reconstructed
  ),
  na.rm = TRUE
)

stopifnot(
  reconstruction_error < 1e-10
)

# Convert the components to long format
air_components_long <- air_components |>
  select(
    Date,
    Observed,
    Trend,
    Seasonal,
    Remainder
  ) |>
  pivot_longer(
    cols = c(
      Observed,
      Trend,
      Seasonal,
      Remainder
    ),
    names_to = "Component_code",
    values_to = "Value"
  ) |>
  mutate(
    Component = factor(
      Component_code,
      levels = c(
        "Observed",
        "Trend",
        "Seasonal",
        "Remainder"
      ),
      labels = c(
        "Observed series: Yt",
        "Trend component: Tt",
        "Seasonal component: St",
        "Remainder component: Rt"
      )
    ),
    Component_code = factor(
      Component_code,
      levels = c(
        "Observed",
        "Trend",
        "Seasonal",
        "Remainder"
      )
    )
  )

# Zero-reference lines for seasonal and remainder panels
zero_lines <- tibble(
  Component = factor(
    c(
      "Seasonal component: St",
      "Remainder component: Rt"
    ),
    levels = levels(
      air_components_long$Component
    )
  ),
  Reference = 0
)

# Robustly identify unusual remainder observations
remainder_center <- median(
  air_components$Remainder,
  na.rm = TRUE
)

remainder_scale <- mad(
  air_components$Remainder,
  center = remainder_center,
  constant = 1.4826,
  na.rm = TRUE
)

# Fallback if MAD is zero or invalid
if (
  !is.finite(remainder_scale) ||
  remainder_scale == 0
) {
  remainder_scale <- sd(
    air_components$Remainder,
    na.rm = TRUE
  )
}

air_components <- air_components |>
  mutate(
    Remainder_robust_z =
      (Remainder - remainder_center) /
      remainder_scale,
    Unusual_remainder =
      abs(Remainder_robust_z) > 2.5
  )

# Data for highlighted remainder observations
unusual_remainders <- air_components |>
  filter(Unusual_remainder) |>
  transmute(
    Date = Date,
    Value = Remainder,
    Component = factor(
      "Remainder component: Rt",
      levels = levels(
        air_components_long$Component
      )
    )
  )

# Select up to two unusual observations for labels
remainder_labels <- unusual_remainders |>
  slice_max(
    order_by = abs(Value),
    n = min(
      2L,
      nrow(unusual_remainders)
    ),
    with_ties = FALSE
  ) |>
  mutate(
    Label = format(
      Date,
      "%b %Y"
    ),
    Label_y = if_else(
      Value >= 0,
      Value + 0.025,
      Value - 0.025
    ),
    Vjust = if_else(
      Value >= 0,
      0,
      1
    )
  )

# Colors for the four components
component_colors <- c(
  "Observed" = "#173F73",
  "Trend" = "#059669",
  "Seasonal" = "#7C3AED",
  "Remainder" = "#64748B"
)

# Build the decomposition figure
decomposition_plot <- ggplot(
  data = air_components_long,
  aes(
    x = Date,
    y = Value,
    color = Component_code,
    group = Component
  )
) +
  geom_hline(
    data = zero_lines,
    aes(
      yintercept = Reference
    ),
    inherit.aes = FALSE,
    color = "#CBD5E1",
    linewidth = 0.55
  ) +
  geom_line(
    linewidth = 0.9,
    lineend = "round"
  ) +
  geom_point(
    data = unusual_remainders,
    aes(
      x = Date,
      y = Value
    ),
    inherit.aes = FALSE,
    color = "#DC2626",
    fill = "white",
    shape = 21,
    size = 3.2,
    stroke = 1.2
  ) +
  geom_text(
    data = remainder_labels,
    aes(
      x = Date,
      y = Label_y,
      label = Label,
      vjust = Vjust
    ),
    inherit.aes = FALSE,
    color = "#B91C1C",
    size = 2.7,
    fontface = "bold",
    check_overlap = TRUE
  ) +
  facet_grid(
    rows = vars(Component),
    scales = "free_y",
    switch = "y",
    labeller = label_value
  ) +
  scale_color_manual(
    values = component_colors,
    guide = "none"
  ) +
  scale_x_date(
    date_breaks = "2 years",
    date_minor_breaks = "1 year",
    date_labels = "%Y",
    expand = expansion(
      mult = c(0.01, 0.025)
    )
  ) +
  scale_y_continuous(
    labels = label_number(
      accuracy = 0.01
    ),
    expand = expansion(
      mult = c(0.15, 0.18)
    )
  ) +
  labs(
    title = "Decomposing the AirPassengers Time Series",
    subtitle = paste0(
      "Additive STL decomposition on the log scale: ",
      "Yt = Tt + St + Rt"
    ),
    x = "Year",
    y = NULL,
    caption = paste0(
      "The logarithmic transformation stabilizes the increasing ",
      "seasonal amplitude. Red circles identify unusually large ",
      "remainder values using a robust standardized threshold."
    )
  ) +
  theme_minimal(
    base_size = 13
  ) +
  theme(
    plot.title = element_text(
      size = 20,
      face = "bold",
      color = "#172033",
      margin = margin(b = 4)
    ),
    plot.subtitle = element_text(
      size = 12.5,
      color = "#475569",
      margin = margin(b = 12)
    ),
    plot.caption = element_text(
      size = 8.5,
      color = "#64748B",
      hjust = 0,
      lineheight = 1.15,
      margin = margin(t = 10)
    ),
    strip.placement = "outside",
    strip.background = element_rect(
      fill = "#F8FAFC",
      color = "#CBD5E1",
      linewidth = 0.5
    ),
    strip.text.y.left = element_text(
      angle = 0,
      size = 9.5,
      face = "bold",
      color = "#334155",
      hjust = 0.5,
      margin = margin(
        t = 6,
        r = 9,
        b = 6,
        l = 9
      )
    ),
    axis.title.x = element_text(
      size = 11,
      face = "bold",
      color = "#334155",
      margin = margin(t = 8)
    ),
    axis.text = element_text(
      size = 8.5,
      color = "#64748B"
    ),
    axis.ticks = element_line(
      color = "#CBD5E1",
      linewidth = 0.4
    ),
    panel.grid.major.x = element_line(
      color = "#E5E7EB",
      linewidth = 0.4
    ),
    panel.grid.major.y = element_line(
      color = "#F1F5F9",
      linewidth = 0.35
    ),
    panel.grid.minor.x = element_line(
      color = "#F1F5F9",
      linewidth = 0.25
    ),
    panel.grid.minor.y = element_blank(),
    panel.spacing.y = grid::unit(
      0.45,
      "lines"
    ),
    plot.margin = margin(
      t = 10,
      r = 24,
      b = 8,
      l = 10
    )
  )

decomposition_plot
```


::: {.notes}

A trend exists when there is a long-term increase or decrease in the data. It does not have to be linear. Sometimes we will refer to a trend as “changing direction”, when it might go from an increasing trend to a decreasing trend.

A seasonal pattern occurs when a time series is affected by seasonal factors such as the time of the year or the day of the week. Seasonality is always of a fixed and known period.

A cycle occurs when the data exhibit rises and falls that are not of a fixed frequency. These fluctuations are usually due to economic conditions, and are often related to the “business cycle”. The duration of these fluctuations is usually at least 2 years.

:::

## Time Series Decomposition

```{r}
library(dplyr)
library(ggplot2)
library(scales)

# ================================================================
# 1. VALIDATE THE MONTHLY DATASET
# ================================================================

required_columns <- c(
  "ticker",
  "month",
  "monthly_adjusted_price"
)

missing_columns <- setdiff(
  required_columns,
  names(monthly_data)
)

if (length(missing_columns) > 0L) {
  stop(
    paste0(
      "The following columns are missing from monthly_data: ",
      paste(
        missing_columns,
        collapse = ", "
      )
    )
  )
}

# ================================================================
# 2. SELECT THE FIRST VALID TICKER
# ================================================================

ticker_selected <- monthly_data |>
  filter(
    !is.na(ticker),
    as.character(ticker) != ""
  ) |>
  distinct(ticker) |>
  slice_head(n = 1L) |>
  pull(ticker) |>
  as.character()

if (
  length(ticker_selected) == 0L ||
  is.na(ticker_selected)
) {
  stop(
    "No valid ticker was found in monthly_data."
  )
}

# ================================================================
# 3. PREPARE THE MONTHLY ADJUSTED-PRICE SERIES
# ================================================================

monthly_trend_data <- monthly_data |>
  filter(
    ticker == ticker_selected
  ) |>
  transmute(
    ticker = as.character(ticker),

    month = as.Date(month),

    monthly_adjusted_price =
      suppressWarnings(
        as.numeric(
          monthly_adjusted_price
        )
      )
  ) |>
  filter(
    !is.na(month),
    is.finite(monthly_adjusted_price)
  ) |>
  group_by(
    ticker,
    month
  ) |>
  summarise(
    monthly_adjusted_price = last(
      monthly_adjusted_price
    ),

    .groups = "drop"
  ) |>
  arrange(month)

n_months <- nrow(
  monthly_trend_data
)

if (n_months < 13L) {
  stop(
    paste0(
      "At least 13 monthly prices are required. ",
      "Only ",
      n_months,
      " were found for ",
      ticker_selected,
      "."
    )
  )
}

# ================================================================
# 4. VERIFY THAT THE SERIES HAS NO MISSING CALENDAR MONTHS
#
# Classical decomposition assumes equally spaced observations.
# ================================================================

expected_months <- seq.Date(
  from = min(
    monthly_trend_data$month
  ),

  to = max(
    monthly_trend_data$month
  ),

  by = "month"
)

observed_months <- monthly_trend_data$month

missing_months <- expected_months[
  !expected_months %in%
    observed_months
]

if (length(missing_months) > 0L) {
  stop(
    paste0(
      "The monthly series contains missing calendar months: ",
      paste(
        format(
          missing_months,
          "%b %Y"
        ),
        collapse = ", "
      ),
      ". Treat the missing months before estimating the trend."
    )
  )
}

# ================================================================
# 5. DEFINE THE CENTERED 2 x 12 MOVING-AVERAGE WEIGHTS
#
# For monthly data, m = 12.
#
# Because 12 is even, the centered moving average is equivalent
# to a 13-term weighted moving average:
#
# 1/24, 1/12, ..., 1/12, 1/24
# ================================================================

trend_weights <- c(
  1 / 24,

  rep(
    1 / 12,
    11L
  ),

  1 / 24
)

stopifnot(
  length(trend_weights) == 13L,

  isTRUE(
    all.equal(
      sum(trend_weights),
      1
    )
  )
)

# ================================================================
# 6. ESTIMATE THE CLASSICAL MONTHLY TREND
#
# T_hat(t) =
# Y_(t-6)/24
# + sum[Y_(t+j)/12], j = -5,...,5
# + Y_(t+6)/24
# ================================================================

monthly_trend_data <- monthly_trend_data |>
  mutate(
    classical_trend = as.numeric(
      stats::filter(
        x = monthly_adjusted_price,

        filter = trend_weights,

        sides = 2
      )
    )
  )

# ================================================================
# 7. RETAIN VALID TREND ESTIMATES
#
# Six observations are lost at each edge because the estimate
# requires six months before and six months after time t.
# ================================================================

valid_trend_data <- monthly_trend_data |>
  filter(
    !is.na(classical_trend),
    is.finite(classical_trend)
  )

expected_trend_count <-
  n_months - 12L

if (
  nrow(valid_trend_data) !=
    expected_trend_count
) {
  stop(
    paste0(
      "Unexpected number of trend estimates. Expected ",
      expected_trend_count,
      " but obtained ",
      nrow(valid_trend_data),
      "."
    )
  )
}

# ================================================================
# 8. SELECT ONE COMPLETE 13-MONTH WINDOW FOR ILLUSTRATION
# ================================================================

valid_indices <- which(
  is.finite(
    monthly_trend_data$classical_trend
  )
)

example_center_index <- valid_indices[
  ceiling(
    length(valid_indices) / 2
  )
]

example_window_indices <- seq.int(
  from = example_center_index - 6L,

  to = example_center_index + 6L
)

example_window <- monthly_trend_data[
  example_window_indices,
  ,
  drop = FALSE
]

example_center <- monthly_trend_data[
  example_center_index,
  ,
  drop = FALSE
]

# ================================================================
# 9. VALIDATE THE EXAMPLE MANUALLY
# ================================================================

manual_example_trend <- sum(
  example_window$monthly_adjusted_price *
    trend_weights
)

filter_example_trend <-
  example_center$classical_trend[1]

if (
  !isTRUE(
    all.equal(
      manual_example_trend,
      filter_example_trend,
      tolerance = 1e-10
    )
  )
) {
  stop(
    "The manual trend estimate does not match the filtered estimate."
  )
}

# ================================================================
# 10. PREPARE THE HIGHLIGHTED WINDOW
# ================================================================

window_region <- data.frame(
  xmin = min(
    example_window$month
  ),

  xmax = max(
    example_window$month
  )
)

# ================================================================
# 11. CREATE FORMATTING AND LABELS
# ================================================================

median_price <- median(
  monthly_trend_data$monthly_adjusted_price,
  na.rm = TRUE
)

price_accuracy <- if (
  median_price >= 100
) {
  1
} else if (
  median_price >= 10
) {
  0.1
} else {
  0.01
}

format_price <- label_number(
  accuracy = price_accuracy,
  big.mark = ","
)

price_range <- diff(
  range(
    monthly_trend_data$monthly_adjusted_price,
    na.rm = TRUE
  )
)

if (
  !is.finite(price_range) ||
  price_range <= 0
) {
  price_range <- 1
}

label_y <-
  filter_example_trend +
  0.15 * price_range

trend_label <- paste0(
  "Centered 2 × 12 moving average\n",
  format(
    example_center$month[1],
    "%b %Y"
  ),
  "\n",
  "Estimated trend = ",
  format_price(
    filter_example_trend
  )
)

# ================================================================
# 12. CREATE THE FIGURE
# ================================================================

monthly_trend_plot <- ggplot(
  data = monthly_trend_data,

  aes(
    x = month
  )
) +

  # Highlight one centered 13-month window

  geom_rect(
    data = window_region,

    aes(
      xmin = xmin,
      xmax = xmax,
      ymin = -Inf,
      ymax = Inf
    ),

    inherit.aes = FALSE,

    fill = "#F59E0B",

    alpha = 0.09
  ) +

  # Monthly adjusted-price series

  geom_line(
    aes(
      y = monthly_adjusted_price,
      color = "Monthly adjusted price"
    ),

    linewidth = 0.65,

    alpha = 0.60,

    lineend = "round"
  ) +

  geom_point(
    aes(
      y = monthly_adjusted_price,
      color = "Monthly adjusted price"
    ),

    size = 1.5,

    alpha = 0.55
  ) +

  # Classical trend estimate

  geom_line(
    data = valid_trend_data,

    aes(
      x = month,
      y = classical_trend,
      color = "Classical trend"
    ),

    inherit.aes = FALSE,

    linewidth = 1.35,

    lineend = "round"
  ) +

  # Prices entering the selected centered window

  geom_point(
    data = example_window,

    aes(
      x = month,
      y = monthly_adjusted_price
    ),

    inherit.aes = FALSE,

    color = "#D97706",

    fill = "white",

    shape = 21,

    size = 3,

    stroke = 0.95
  ) +

  # Estimated trend at the center of the selected window

  geom_point(
    data = example_center,

    aes(
      x = month,
      y = classical_trend
    ),

    inherit.aes = FALSE,

    color = "#047857",

    fill = "white",

    shape = 23,

    size = 5,

    stroke = 1.25
  ) +

  # Label for the example estimate

  annotate(
    geom = "label",

    x = example_center$month[1],

    y = label_y,

    label = trend_label,

    color = "#047857",

    fill = "white",

    fontface = "bold",

    size = 3.1,

    lineheight = 1,

    label.size = 0.3
  ) +

  # Boundaries of the region with valid centered estimates

  geom_vline(
    xintercept = as.numeric(
      min(
        valid_trend_data$month
      )
    ),

    color = "#94A3B8",

    linewidth = 0.5,

    linetype = "dotted"
  ) +

  geom_vline(
    xintercept = as.numeric(
      max(
        valid_trend_data$month
      )
    ),

    color = "#94A3B8",

    linewidth = 0.5,

    linetype = "dotted"
  ) +

  # Line colors

  scale_color_manual(
    name = NULL,

    values = c(
      "Monthly adjusted price" =
        "#94A3B8",

      "Classical trend" =
        "#059669"
    ),

    breaks = c(
      "Monthly adjusted price",
      "Classical trend"
    )
  ) +

  # Horizontal axis

  scale_x_date(
    date_breaks = "1 year",

    date_minor_breaks = "3 months",

    date_labels = "%Y",

    expand = expansion(
      mult = c(
        0.01,
        0.02
      )
    )
  ) +

  # Vertical axis

  scale_y_continuous(
    labels = format_price,

    expand = expansion(
      mult = c(
        0.07,
        0.22
      )
    )
  ) +

  # Titles

  labs(
    title =
      "How Is the Monthly Trend Estimated?",

    subtitle = paste0(
      ticker_selected,
      " monthly adjusted prices | ",
      "Classical centered 2 × 12 moving average"
    ),

    x = "Month",

    y = "Monthly adjusted price",

    caption = paste0(
      "The gray series contains the last valid adjusted price ",
      "observed in each month. The green line is the classical ",
      "trend estimate. The highlighted region shows the 13 prices ",
      "used in one centered 2 × 12 moving-average calculation."
    )
  ) +

  # Presentation theme

  theme_minimal(
    base_size = 13
  ) +

  theme(
    plot.title = element_text(
      size = 20,
      face = "bold",
      color = "#172033",
      margin = margin(
        b = 4
      )
    ),

    plot.subtitle = element_text(
      size = 11.5,
      color = "#475569",
      margin = margin(
        b = 13
      )
    ),

    plot.caption = element_text(
      size = 8.5,
      color = "#64748B",
      hjust = 0,
      lineheight = 1.15,
      margin = margin(
        t = 10
      )
    ),

    axis.title = element_text(
      size = 11,
      face = "bold",
      color = "#334155"
    ),

    axis.text = element_text(
      size = 8.5,
      color = "#64748B"
    ),

    panel.grid.major = element_line(
      color = "#E5E7EB",
      linewidth = 0.4
    ),

    panel.grid.minor.x = element_line(
      color = "#F1F5F9",
      linewidth = 0.25
    ),

    panel.grid.minor.y = element_blank(),

    legend.position = "bottom",

    legend.text = element_text(
      size = 9.5,
      color = "#475569"
    ),

    legend.key.width = grid::unit(
      1.7,
      "cm"
    ),

    plot.margin = margin(
      t = 10,
      r = 20,
      b = 8,
      l = 10
    )
  )

# ================================================================
# 13. DISPLAY THE FIGURE
# ================================================================

monthly_trend_plot
```

## Time Series Decomposition

```{r}
# ================================================================
# DETRENDED MONTHLY ADJUSTED-PRICE SERIES
# Classical decomposition with a centered 2 x 12 moving average
# ================================================================

library(dplyr)
library(ggplot2)
library(scales)

# ================================================================
# 1. VALIDATE THE INPUT DATASET
#
# This code assumes that monthly_data was created previously
# and contains one or more tickers with monthly adjusted prices.
# ================================================================

if (!exists("monthly_data")) {
  stop(
    paste0(
      "The object 'monthly_data' does not exist. ",
      "Create the monthly adjusted-price dataset before running ",
      "this chunk."
    )
  )
}

required_columns <- c(
  "ticker",
  "month",
  "monthly_adjusted_price"
)

missing_columns <- setdiff(
  required_columns,
  names(monthly_data)
)

if (length(missing_columns) > 0L) {
  stop(
    paste0(
      "The following required columns are missing from monthly_data: ",
      paste(
        missing_columns,
        collapse = ", "
      )
    )
  )
}

# ================================================================
# 2. SELECT THE FIRST VALID TICKER
# ================================================================

available_tickers <- monthly_data |>
  transmute(
    ticker = as.character(ticker)
  ) |>
  filter(
    !is.na(ticker),
    ticker != ""
  ) |>
  distinct(ticker) |>
  pull(ticker)

if (length(available_tickers) == 0L) {
  stop(
    "No valid ticker was found in monthly_data."
  )
}

ticker_selected <- available_tickers[1]

# ================================================================
# 3. PREPARE THE MONTHLY ADJUSTED-PRICE SERIES
#
# The code guarantees one price for each ticker-month pair.
# If duplicates exist, the final finite value is retained.
# ================================================================

monthly_price_series <- monthly_data |>
  transmute(
    ticker = as.character(ticker),

    month = as.Date(month),

    monthly_adjusted_price = suppressWarnings(
      as.numeric(
        monthly_adjusted_price
      )
    )
  ) |>
  filter(
    ticker == ticker_selected,
    !is.na(month),
    is.finite(monthly_adjusted_price)
  ) |>
  arrange(
    month
  ) |>
  group_by(
    ticker,
    month
  ) |>
  summarise(
    monthly_adjusted_price = last(
      monthly_adjusted_price
    ),

    .groups = "drop"
  ) |>
  arrange(
    month
  )

n_months <- nrow(
  monthly_price_series
)

if (n_months < 13L) {
  stop(
    paste0(
      "At least 13 monthly observations are required to calculate ",
      "a centered 2 x 12 moving average. Only ",
      n_months,
      " were found for ",
      ticker_selected,
      "."
    )
  )
}

# ================================================================
# 4. CHECK THE MONTHLY FREQUENCY
#
# Classical decomposition assumes an equally spaced monthly series.
# ================================================================

expected_month_sequence <- seq.Date(
  from = min(
    monthly_price_series$month
  ),

  to = max(
    monthly_price_series$month
  ),

  by = "month"
)

missing_months <- expected_month_sequence[
  !expected_month_sequence %in%
    monthly_price_series$month
]

if (length(missing_months) > 0L) {
  stop(
    paste0(
      "The series contains missing calendar months: ",
      paste(
        format(
          missing_months,
          "%b %Y"
        ),
        collapse = ", "
      ),
      ". Classical decomposition requires an equally spaced series."
    )
  )
}

# ================================================================
# 5. DEFINE THE CENTERED 2 x 12 MOVING-AVERAGE WEIGHTS
#
# With monthly data, the seasonal period is m = 12.
#
# Since m is even, the centered trend is obtained from a
# 13-observation weighted moving average:
#
#            1      5             1
# T_hat_t = ---Y_t-6 + sum ---Y_t+j + ---Y_t+6
#           24     j=-5 12        24
# ================================================================

trend_weights <- c(
  1 / 24,

  rep(
    1 / 12,
    times = 11L
  ),

  1 / 24
)

if (length(trend_weights) != 13L) {
  stop(
    "The centered moving-average filter must contain 13 weights."
  )
}

if (
  !isTRUE(
    all.equal(
      sum(trend_weights),
      1,
      tolerance = 1e-12
    )
  )
) {
  stop(
    "The centered moving-average weights do not sum to one."
  )
}

# ================================================================
# 6. ESTIMATE THE CLASSICAL TREND
#
# stats::filter(..., sides = 2) applies the weights symmetrically.
# Six months before and six months after t are required.
# ================================================================

estimated_trend <- as.numeric(
  stats::filter(
    x = monthly_price_series$monthly_adjusted_price,

    filter = trend_weights,

    sides = 2
  )
)

monthly_price_series <- monthly_price_series |>
  mutate(
    estimated_trend = estimated_trend
  )

# ================================================================
# 7. CALCULATE THE DETRENDED SERIES
#
# D_t = Y_t - T_hat_t
#
# For the classical additive decomposition:
#
# D_t = S_t + R_t
# ================================================================

monthly_price_series <- monthly_price_series |>
  mutate(
    detrended_price =
      monthly_adjusted_price -
      estimated_trend
  )

# ================================================================
# 8. RETAIN DATES WITH VALID CENTERED TREND ESTIMATES
#
# The first and final six months do not have trend estimates.
# ================================================================

detrended_data <- monthly_price_series |>
  filter(
    is.finite(estimated_trend),
    is.finite(detrended_price)
  )

n_detrended_observations <- nrow(
  detrended_data
)

expected_detrended_observations <-
  n_months - 12L

if (
  n_detrended_observations !=
    expected_detrended_observations
) {
  stop(
    paste0(
      "Unexpected number of detrended observations. Expected ",
      expected_detrended_observations,
      " but obtained ",
      n_detrended_observations,
      "."
    )
  )
}

# ================================================================
# 9. VALIDATE THE DETRENDING IDENTITY
#
# Wherever the trend is available:
#
# Y_t = T_hat_t + D_t
# ================================================================

reconstruction_error <- max(
  abs(
    detrended_data$monthly_adjusted_price -
      (
        detrended_data$estimated_trend +
          detrended_data$detrended_price
      )
  ),
  na.rm = TRUE
)

if (
  !is.finite(reconstruction_error) ||
  reconstruction_error > 1e-10
) {
  stop(
    paste0(
      "The detrending identity failed. Maximum reconstruction error: ",
      format(
        reconstruction_error,
        scientific = TRUE
      )
    )
  )
}

# ================================================================
# 10. CLASSIFY VALUES RELATIVE TO THE TREND
# ================================================================

detrended_data <- detrended_data |>
  mutate(
    position_relative_to_trend = case_when(
      detrended_price > 0 ~
        "Price above trend",

      detrended_price < 0 ~
        "Price below trend",

      TRUE ~
        "Price equal to trend"
    )
  )

# ================================================================
# 11. CALCULATE DESCRIPTIVE STATISTICS
# ================================================================

mean_detrended <- mean(
  detrended_data$detrended_price,
  na.rm = TRUE
)

sd_detrended <- sd(
  detrended_data$detrended_price,
  na.rm = TRUE
)

maximum_detrended <- max(
  detrended_data$detrended_price,
  na.rm = TRUE
)

minimum_detrended <- min(
  detrended_data$detrended_price,
  na.rm = TRUE
)

# ================================================================
# 12. IDENTIFY LARGE ABSOLUTE DEVIATIONS
#
# The largest 5% are emphasized only as a visualization device.
# This is not presented as a formal outlier-detection rule.
# ================================================================

large_deviation_threshold <- quantile(
  abs(
    detrended_data$detrended_price
  ),

  probs = 0.95,

  names = FALSE,

  na.rm = TRUE,

  type = 7
)

detrended_data <- detrended_data |>
  mutate(
    deviation_class = if_else(
      abs(detrended_price) >=
        large_deviation_threshold,

      "Large absolute deviation",

      "Regular deviation"
    )
  )

regular_deviations <- detrended_data |>
  filter(
    deviation_class ==
      "Regular deviation"
  )

large_deviations <- detrended_data |>
  filter(
    deviation_class ==
      "Large absolute deviation"
  )

# ================================================================
# 13. IDENTIFY THE LARGEST POSITIVE AND NEGATIVE DEVIATIONS
# ================================================================

largest_positive_deviation <- detrended_data |>
  slice_max(
    order_by = detrended_price,

    n = 1L,

    with_ties = FALSE
  )

largest_negative_deviation <- detrended_data |>
  slice_min(
    order_by = detrended_price,

    n = 1L,

    with_ties = FALSE
  )

extreme_deviations <- bind_rows(
  largest_positive_deviation,
  largest_negative_deviation
) |>
  distinct(
    month,

    .keep_all = TRUE
  )

# ================================================================
# 14. DEFINE THE NUMBER FORMAT
#
# Detrended values retain the price unit.
# ================================================================

median_price <- median(
  detrended_data$monthly_adjusted_price,

  na.rm = TRUE
)

price_accuracy <- if (
  median_price >= 100
) {
  1
} else if (
  median_price >= 10
) {
  0.1
} else {
  0.01
}

price_formatter <- label_number(
  accuracy = price_accuracy,

  big.mark = ","
)

# ================================================================
# 15. CREATE LABELS FOR EXTREME DEVIATIONS
# ================================================================

extreme_deviations <- extreme_deviations |>
  mutate(
    extreme_label = paste0(
      format(
        month,
        "%b %Y"
      ),
      "\n",
      if_else(
        detrended_price > 0,
        "Above trend: ",
        "Below trend: "
      ),
      price_formatter(
        detrended_price
      )
    ),

    vertical_justification = if_else(
      detrended_price >= 0,

      -0.75,

      1.65
    )
  )

# ================================================================
# 16. CREATE A SYMMETRIC VERTICAL SCALE
#
# A symmetric scale makes positive and negative deviations directly
# comparable.
# ================================================================

maximum_absolute_deviation <- max(
  abs(
    detrended_data$detrended_price
  ),

  na.rm = TRUE
)

if (
  !is.finite(maximum_absolute_deviation) ||
  maximum_absolute_deviation <= 0
) {
  maximum_absolute_deviation <- 1
}

vertical_limit <-
  1.22 *
  maximum_absolute_deviation

# ================================================================
# 17. CREATE LABEL POSITIONS FOR EXPLANATORY TEXT
# ================================================================

first_valid_month <- min(
  detrended_data$month
)

last_valid_month <- max(
  detrended_data$month
)

positive_region_y <-
  0.78 *
  vertical_limit

negative_region_y <-
  -0.78 *
  vertical_limit

# ================================================================
# 18. CREATE THE DETRENDED-SERIES FIGURE
# ================================================================

monthly_detrended_plot <- ggplot(
  data = detrended_data,

  aes(
    x = month,
    y = detrended_price
  )
) +

  # --------------------------------------------------------------
  # Positive-deviation area
  # --------------------------------------------------------------

  geom_ribbon(
    aes(
      ymin = 0,

      ymax = ifelse(
        detrended_price > 0,
        detrended_price,
        0
      )
    ),

    fill = "#2F6FA5",

    alpha = 0.15
  ) +

  # --------------------------------------------------------------
  # Negative-deviation area
  # --------------------------------------------------------------

  geom_ribbon(
    aes(
      ymin = ifelse(
        detrended_price < 0,
        detrended_price,
        0
      ),

      ymax = 0
    ),

    fill = "#DC2626",

    alpha = 0.15
  ) +

  # --------------------------------------------------------------
  # Zero-reference line
  #
  # At zero:
  #
  # observed price = estimated trend
  # --------------------------------------------------------------

  geom_hline(
    yintercept = 0,

    color = "#334155",

    linewidth = 0.8
  ) +

  # --------------------------------------------------------------
  # Complete detrended series
  # --------------------------------------------------------------

  geom_line(
    color = "#64748B",

    linewidth = 0.7,

    alpha = 0.80,

    lineend = "round"
  ) +

  # --------------------------------------------------------------
  # Regular observations
  # --------------------------------------------------------------

  geom_point(
    data = regular_deviations,

    aes(
      x = month,
      y = detrended_price,
      color = position_relative_to_trend
    ),

    inherit.aes = FALSE,

    size = 2,

    alpha = 0.62
  ) +

  # --------------------------------------------------------------
  # Large absolute deviations
  # --------------------------------------------------------------

  geom_point(
    data = large_deviations,

    aes(
      x = month,
      y = detrended_price
    ),

    inherit.aes = FALSE,

    color = "#D97706",

    fill = "white",

    shape = 21,

    size = 3.5,

    stroke = 1.1
  ) +

  # --------------------------------------------------------------
  # Label the largest positive and negative deviations
  # --------------------------------------------------------------

  geom_text(
    data = extreme_deviations,

    aes(
      x = month,
      y = detrended_price,
      label = extreme_label,
      vjust = vertical_justification
    ),

    inherit.aes = FALSE,

    color = "#334155",

    size = 2.9,

    fontface = "bold",

    lineheight = 0.95,

    check_overlap = TRUE
  ) +

  # --------------------------------------------------------------
  # Explain the positive region
  # --------------------------------------------------------------

  annotate(
    geom = "label",

    x = first_valid_month,

    y = positive_region_y,

    label = paste0(
      "D(t) > 0\n",
      "Price above trend"
    ),

    hjust = 0,

    color = "#1D4ED8",

    fill = "white",

    fontface = "bold",

    size = 3,

    lineheight = 1,

    label.size = 0.25
  ) +

  # --------------------------------------------------------------
  # Explain the negative region
  # --------------------------------------------------------------

  annotate(
    geom = "label",

    x = first_valid_month,

    y = negative_region_y,

    label = paste0(
      "D(t) < 0\n",
      "Price below trend"
    ),

    hjust = 0,

    color = "#B91C1C",

    fill = "white",

    fontface = "bold",

    size = 3,

    lineheight = 1,

    label.size = 0.25
  ) +

  # --------------------------------------------------------------
  # Colors
  # --------------------------------------------------------------

  scale_color_manual(
    name = NULL,

    values = c(
      "Price above trend" =
        "#2F6FA5",

      "Price below trend" =
        "#DC2626",

      "Price equal to trend" =
        "#059669"
    ),

    breaks = c(
      "Price above trend",
      "Price below trend",
      "Price equal to trend"
    )
  ) +

  # --------------------------------------------------------------
  # Horizontal scale
  # --------------------------------------------------------------

  scale_x_date(
    date_breaks = "1 year",

    date_minor_breaks = "3 months",

    date_labels = "%Y",

    expand = expansion(
      mult = c(
        0.01,
        0.02
      )
    )
  ) +

  # --------------------------------------------------------------
  # Symmetric vertical scale
  # --------------------------------------------------------------

  scale_y_continuous(
    limits = c(
      -vertical_limit,
      vertical_limit
    ),

    labels = price_formatter,

    expand = expansion(
      mult = c(
        0,
        0
      )
    )
  ) +

  # --------------------------------------------------------------
  # Titles
  # --------------------------------------------------------------

  labs(
    title =
      "What Remains After Removing the Monthly Trend?",

    subtitle = paste0(
      ticker_selected,
      " monthly adjusted prices | ",
      "D(t) = Y(t) - T(t)"
    ),

    x = "Month",

    y = "Detrended price",

    caption = paste0(
      "The trend is estimated using a centered 2 × 12 moving average. ",
      "Blue areas represent prices above the estimated trend; ",
      "red areas represent prices below it. ",
      "Orange circles identify the 5% largest absolute deviations. ",
      "The detrended series contains seasonal and remainder variation."
    )
  ) +

  # --------------------------------------------------------------
  # Presentation theme
  # --------------------------------------------------------------

  theme_minimal(
    base_size = 13
  ) +

  theme(
    plot.title = element_text(
      size = 20,

      face = "bold",

      color = "#172033",

      margin = margin(
        b = 4
      )
    ),

    plot.subtitle = element_text(
      size = 11.5,

      color = "#475569",

      margin = margin(
        b = 13
      )
    ),

    plot.caption = element_text(
      size = 8.5,

      color = "#64748B",

      hjust = 0,

      lineheight = 1.15,

      margin = margin(
        t = 10
      )
    ),

    axis.title = element_text(
      size = 11,

      face = "bold",

      color = "#334155"
    ),

    axis.text = element_text(
      size = 8.5,

      color = "#64748B"
    ),

    panel.grid.major = element_line(
      color = "#E5E7EB",

      linewidth = 0.4
    ),

    panel.grid.minor.x = element_line(
      color = "#F1F5F9",

      linewidth = 0.25
    ),

    panel.grid.minor.y = element_blank(),

    legend.position = "bottom",

    legend.text = element_text(
      size = 9.5,

      color = "#475569"
    ),

    legend.key.width = grid::unit(
      0.9,
      "cm"
    ),

    plot.margin = margin(
      t = 10,
      r = 20,
      b = 8,
      l = 10
    )
  ) +

  guides(
    color = guide_legend(
      override.aes = list(
        size = 3,

        alpha = 0.85
      )
    )
  )

# ================================================================
# 19. DISPLAY THE FIGURE
# ================================================================

monthly_detrended_plot
```

## Time Series Decomposition

```{r}
# ================================================================
# CLASSICAL SEASONAL COMPONENT
# Monthly adjusted prices
# ================================================================

library(dplyr)
library(ggplot2)
library(scales)

# ================================================================
# 1. VALIDATE THE MONTHLY DATA
# ================================================================

if (!exists("monthly_data")) {
  stop(
    paste0(
      "The object 'monthly_data' does not exist. ",
      "Run the daily-to-monthly aggregation code first."
    )
  )
}

required_columns <- c(
  "ticker",
  "month",
  "monthly_adjusted_price"
)

missing_columns <- setdiff(
  required_columns,
  names(monthly_data)
)

if (length(missing_columns) > 0L) {
  stop(
    paste0(
      "The following columns are missing from monthly_data: ",
      paste(
        missing_columns,
        collapse = ", "
      )
    )
  )
}

# ================================================================
# 2. SELECT THE FIRST VALID TICKER
# ================================================================

ticker_selected <- monthly_data |>
  filter(
    !is.na(ticker),
    as.character(ticker) != ""
  ) |>
  distinct(ticker) |>
  slice_head(n = 1L) |>
  pull(ticker) |>
  as.character()

if (
  length(ticker_selected) == 0L ||
  is.na(ticker_selected)
) {
  stop(
    "No valid ticker was found in monthly_data."
  )
}

# ================================================================
# 3. PREPARE ONE PRICE PER CALENDAR MONTH
# ================================================================

monthly_series <- monthly_data |>
  transmute(
    ticker = as.character(ticker),

    month = as.Date(month),

    monthly_adjusted_price = suppressWarnings(
      as.numeric(
        monthly_adjusted_price
      )
    )
  ) |>
  filter(
    ticker == ticker_selected,
    !is.na(month),
    is.finite(monthly_adjusted_price)
  ) |>
  arrange(month) |>
  group_by(
    ticker,
    month
  ) |>
  summarise(
    monthly_adjusted_price = last(
      monthly_adjusted_price
    ),

    .groups = "drop"
  ) |>
  arrange(month)

n_months <- nrow(
  monthly_series
)

if (n_months < 24L) {
  stop(
    paste0(
      "At least 24 monthly observations are required. ",
      "Only ",
      n_months,
      " observations were found for ",
      ticker_selected,
      "."
    )
  )
}

# ================================================================
# 4. CHECK FOR DUPLICATED MONTHS
# ================================================================

duplicated_months <- monthly_series |>
  count(month) |>
  filter(n > 1L)

if (nrow(duplicated_months) > 0L) {
  stop(
    "The monthly series contains duplicated months."
  )
}

# ================================================================
# 5. CHECK FOR MISSING CALENDAR MONTHS
#
# Classical decomposition requires equally spaced observations.
# ================================================================

expected_months <- seq.Date(
  from = min(
    monthly_series$month
  ),

  to = max(
    monthly_series$month
  ),

  by = "month"
)

missing_months <- expected_months[
  !expected_months %in%
    monthly_series$month
]

if (length(missing_months) > 0L) {
  stop(
    paste0(
      "The series contains missing calendar months: ",
      paste(
        format(
          missing_months,
          "%b %Y"
        ),
        collapse = ", "
      ),
      ". Complete or treat the missing months before ",
      "applying classical decomposition."
    )
  )
}

# ================================================================
# 6. CREATE THE MONTHLY TIME-SERIES OBJECT
#
# frequency = 12 defines an annual seasonal cycle.
# ================================================================

start_year <- as.integer(
  format(
    min(monthly_series$month),
    "%Y"
  )
)

start_month <- as.integer(
  format(
    min(monthly_series$month),
    "%m"
  )
)

monthly_price_ts <- ts(
  data = monthly_series$monthly_adjusted_price,

  start = c(
    start_year,
    start_month
  ),

  frequency = 12
)

# ================================================================
# 7. APPLY CLASSICAL ADDITIVE DECOMPOSITION
#
# Y_t = T_t + S_t + R_t
#
# stats::decompose() estimates:
# - trend with a centered moving average;
# - seasonal effects from detrended monthly values;
# - remainder by subtraction.
# ================================================================

classical_decomposition <- stats::decompose(
  x = monthly_price_ts,

  type = "additive"
)

# ================================================================
# 8. EXTRACT THE COMPONENTS
# ================================================================

monthly_series <- monthly_series |>
  mutate(
    observed_price = as.numeric(
      classical_decomposition$x
    ),

    trend_component = as.numeric(
      classical_decomposition$trend
    ),

    seasonal_component = as.numeric(
      classical_decomposition$seasonal
    ),

    remainder_component = as.numeric(
      classical_decomposition$random
    ),

    calendar_month_number = as.integer(
      format(
        month,
        "%m"
      )
    ),

    calendar_month = factor(
      calendar_month_number,

      levels = 1:12,

      labels = month.abb,

      ordered = TRUE
    )
  )

# ================================================================
# 9. VALIDATE THE SEASONAL COMPONENT
# ================================================================

if (
  any(
    !is.finite(
      monthly_series$seasonal_component
    )
  )
) {
  stop(
    "The classical decomposition produced invalid seasonal values."
  )
}

# ================================================================
# 10. EXTRACT THE 12 UNIQUE SEASONAL INDICES
#
# Classical decomposition estimates one effect for each month
# and repeats the same 12-month pattern every year.
# ================================================================

seasonal_indices <- monthly_series |>
  group_by(
    calendar_month_number,
    calendar_month
  ) |>
  summarise(
    seasonal_effect = mean(
      seasonal_component,
      na.rm = TRUE
    ),

    .groups = "drop"
  ) |>
  arrange(
    calendar_month_number
  )

if (nrow(seasonal_indices) != 12L) {
  stop(
    paste0(
      "Expected 12 seasonal indices, but obtained ",
      nrow(seasonal_indices),
      "."
    )
  )
}

# ================================================================
# 11. VALIDATE THE ADDITIVE CENTERING CONSTRAINT
#
# The 12 monthly seasonal effects must sum to approximately zero.
# ================================================================

sum_seasonal_effects <- sum(
  seasonal_indices$seasonal_effect
)

if (
  !isTRUE(
    all.equal(
      sum_seasonal_effects,
      0,
      tolerance = 1e-8
    )
  )
) {
  stop(
    paste0(
      "The seasonal effects do not sum to zero. ",
      "Calculated sum: ",
      format(
        sum_seasonal_effects,
        scientific = TRUE
      )
    )
  )
}

# ================================================================
# 12. VALIDATE THE ADDITIVE DECOMPOSITION
#
# For dates where trend and remainder are available:
#
# Y_t = T_t + S_t + R_t
# ================================================================

complete_components <- monthly_series |>
  filter(
    is.finite(trend_component),
    is.finite(seasonal_component),
    is.finite(remainder_component)
  )

reconstruction_error <- max(
  abs(
    complete_components$observed_price -
      (
        complete_components$trend_component +
        complete_components$seasonal_component +
        complete_components$remainder_component
      )
  ),
  na.rm = TRUE
)

if (
  !is.finite(reconstruction_error) ||
  reconstruction_error > 1e-8
) {
  stop(
    paste0(
      "The additive decomposition identity failed. ",
      "Maximum reconstruction error: ",
      format(
        reconstruction_error,
        scientific = TRUE
      )
    )
  )
}

# ================================================================
# 13. CLASSIFY POSITIVE AND NEGATIVE SEASONAL EFFECTS
# ================================================================

monthly_series <- monthly_series |>
  mutate(
    seasonal_direction = case_when(
      seasonal_component > 0 ~
        "Raises price relative to trend",

      seasonal_component < 0 ~
        "Lowers price relative to trend",

      TRUE ~
        "No seasonal effect"
    )
  )

# ================================================================
# 14. PREPARE POSITIVE AND NEGATIVE AREA VALUES
#
# These columns ensure that geom_ribbon() draws the positive and
# negative regions correctly.
# ================================================================

monthly_series <- monthly_series |>
  mutate(
    positive_seasonal = if_else(
      seasonal_component > 0,

      seasonal_component,

      0
    ),

    negative_seasonal = if_else(
      seasonal_component < 0,

      seasonal_component,

      0
    )
  )

# ================================================================
# 15. IDENTIFY THE LARGEST POSITIVE AND NEGATIVE EFFECTS
#
# The seasonal pattern repeats every year. The final occurrence
# of the highest and lowest seasonal months is highlighted.
# ================================================================

highest_month_number <- seasonal_indices |>
  slice_max(
    order_by = seasonal_effect,

    n = 1L,

    with_ties = FALSE
  ) |>
  pull(
    calendar_month_number
  )

lowest_month_number <- seasonal_indices |>
  slice_min(
    order_by = seasonal_effect,

    n = 1L,

    with_ties = FALSE
  ) |>
  pull(
    calendar_month_number
  )

highest_effect_point <- monthly_series |>
  filter(
    calendar_month_number ==
      highest_month_number
  ) |>
  slice_tail(
    n = 1L
  )

lowest_effect_point <- monthly_series |>
  filter(
    calendar_month_number ==
      lowest_month_number
  ) |>
  slice_tail(
    n = 1L
  )

highlight_points <- bind_rows(
  highest_effect_point,
  lowest_effect_point
)

# ================================================================
# 16. DEFINE THE PRICE FORMAT
#
# Seasonal effects are expressed in the original price unit.
# ================================================================

median_price <- median(
  monthly_series$monthly_adjusted_price,

  na.rm = TRUE
)

price_accuracy <- if (
  median_price >= 100
) {
  1
} else if (
  median_price >= 10
) {
  0.1
} else {
  0.01
}

price_formatter <- label_number(
  accuracy = price_accuracy,

  big.mark = ","
)

# ================================================================
# 17. CREATE HIGHLIGHT LABELS
# ================================================================

highlight_points <- highlight_points |>
  mutate(
    label = paste0(
      as.character(
        calendar_month
      ),
      " seasonal effect\n",
      if_else(
        seasonal_component > 0,
        "+",
        ""
      ),
      price_formatter(
        seasonal_component
      )
    ),

    vertical_justification = if_else(
      seasonal_component > 0,

      -0.8,

      1.8
    )
  )

# ================================================================
# 18. DEFINE A SYMMETRIC VERTICAL SCALE
# ================================================================

maximum_absolute_effect <- max(
  abs(
    monthly_series$seasonal_component
  ),

  na.rm = TRUE
)

if (
  !is.finite(maximum_absolute_effect) ||
  maximum_absolute_effect <= 0
) {
  maximum_absolute_effect <- 1
}

vertical_limit <- 1.30 *
  maximum_absolute_effect

# ================================================================
# 19. CREATE THE SEASONAL-COMPONENT TIME-SERIES PLOT
# ================================================================

seasonal_component_plot <- ggplot(
  data = monthly_series,

  aes(
    x = month,
    y = seasonal_component
  )
) +

  # --------------------------------------------------------------
  # Positive seasonal region
  # --------------------------------------------------------------

  geom_ribbon(
    aes(
      ymin = 0,
      ymax = positive_seasonal
    ),

    fill = "#2F6FA5",

    alpha = 0.17
  ) +

  # --------------------------------------------------------------
  # Negative seasonal region
  # --------------------------------------------------------------

  geom_ribbon(
    aes(
      ymin = negative_seasonal,
      ymax = 0
    ),

    fill = "#DC2626",

    alpha = 0.17
  ) +

  # --------------------------------------------------------------
  # Zero reference
  # --------------------------------------------------------------

  geom_hline(
    yintercept = 0,

    color = "#334155",

    linewidth = 0.8
  ) +

  # --------------------------------------------------------------
  # Seasonal component through time
  # --------------------------------------------------------------

  geom_line(
    color = "#64748B",

    linewidth = 0.75,

    alpha = 0.88,

    lineend = "round"
  ) +

  # --------------------------------------------------------------
  # Monthly seasonal observations
  # --------------------------------------------------------------

  geom_point(
    aes(
      color = seasonal_direction
    ),

    size = 2,

    alpha = 0.72
  ) +

  # --------------------------------------------------------------
  # Highlight the strongest positive and negative months
  # --------------------------------------------------------------

  geom_point(
    data = highlight_points,

    aes(
      x = month,
      y = seasonal_component
    ),

    inherit.aes = FALSE,

    color = "#D97706",

    fill = "white",

    shape = 21,

    size = 4,

    stroke = 1.15
  ) +

  # --------------------------------------------------------------
  # Label the highlighted seasonal effects
  # --------------------------------------------------------------

  geom_text(
    data = highlight_points,

    aes(
      x = month,
      y = seasonal_component,
      label = label,
      vjust = vertical_justification
    ),

    inherit.aes = FALSE,

    color = "#334155",

    size = 3,

    fontface = "bold",

    lineheight = 0.95,

    check_overlap = TRUE
  ) +

  # --------------------------------------------------------------
  # Point colors
  # --------------------------------------------------------------

  scale_color_manual(
    name = NULL,

    values = c(
      "Raises price relative to trend" =
        "#2F6FA5",

      "Lowers price relative to trend" =
        "#DC2626",

      "No seasonal effect" =
        "#059669"
    ),

    breaks = c(
      "Raises price relative to trend",
      "Lowers price relative to trend",
      "No seasonal effect"
    )
  ) +

  # --------------------------------------------------------------
  # Time axis
  # --------------------------------------------------------------

  scale_x_date(
    date_breaks = "1 year",

    date_minor_breaks = "3 months",

    date_labels = "%Y",

    expand = expansion(
      mult = c(
        0.01,
        0.02
      )
    )
  ) +

  # --------------------------------------------------------------
  # Seasonal-effect axis
  # --------------------------------------------------------------

  scale_y_continuous(
    limits = c(
      -vertical_limit,
      vertical_limit
    ),

    labels = price_formatter,

    breaks = pretty_breaks(
      n = 5
    ),

    expand = expansion(
      mult = c(
        0,
        0
      )
    )
  ) +

  # --------------------------------------------------------------
  # Titles
  # --------------------------------------------------------------

  labs(
    title =
      "How Does Seasonality Raise or Lower Monthly Price?",

    subtitle = paste0(
      ticker_selected,
      " monthly adjusted prices | ",
      "Classical additive seasonal component"
    ),

    x = "Month",

    y = expression(
      hat(S)[t]
    ),

    caption = paste0(
      "The seasonal component repeats the same 12 monthly effects ",
      "in each year. Blue values raise price relative to the trend; ",
      "red values lower it. The 12 unique seasonal effects sum to zero."
    )
  ) +

  # --------------------------------------------------------------
  # Presentation theme
  # --------------------------------------------------------------

  theme_minimal(
    base_size = 13
  ) +

  theme(
    plot.title = element_text(
      size = 20,

      face = "bold",

      color = "#172033",

      margin = margin(
        b = 4
      )
    ),

    plot.subtitle = element_text(
      size = 11.5,

      color = "#475569",

      margin = margin(
        b = 13
      )
    ),

    plot.caption = element_text(
      size = 8.5,

      color = "#64748B",

      hjust = 0,

      lineheight = 1.15,

      margin = margin(
        t = 10
      )
    ),

    axis.title = element_text(
      size = 11,

      face = "bold",

      color = "#334155"
    ),

    axis.text = element_text(
      size = 8.5,

      color = "#64748B"
    ),

    panel.grid.major = element_line(
      color = "#E5E7EB",

      linewidth = 0.4
    ),

    panel.grid.minor.x = element_line(
      color = "#F1F5F9",

      linewidth = 0.25
    ),

    panel.grid.minor.y = element_blank(),

    legend.position = "bottom",

    legend.text = element_text(
      size = 9.5,

      color = "#475569"
    ),

    legend.key.width = grid::unit(
      0.9,
      "cm"
    ),

    plot.margin = margin(
      t = 10,
      r = 20,
      b = 8,
      l = 10
    )
  ) +

  guides(
    color = guide_legend(
      override.aes = list(
        size = 3,

        alpha = 0.9
      )
    )
  )

# ================================================================
# 20. DISPLAY THE FIGURE
# ================================================================

seasonal_component_plot
```

## Time Series Decomposition

```{r}
#| label: seasonally-adjusted-monthly-price
#| echo: false
#| warning: false
#| message: false
#| fig-width: 12
#| fig-height: 6.5
#| dpi: 300

# ================================================================
# SEASONALLY ADJUSTED MONTHLY PRICE SERIES
# Classical additive decomposition
# ================================================================

library(dplyr)
library(ggplot2)
library(scales)

# ================================================================
# 1. VALIDATE THE INPUT DATASET
# ================================================================

if (!exists("monthly_data")) {

  stop(
    paste0(
      "The object 'monthly_data' does not exist. ",
      "Create the monthly adjusted-price dataset before ",
      "running this chunk."
    )
  )
}

required_columns <- c(
  "ticker",
  "month",
  "monthly_adjusted_price"
)

missing_columns <- setdiff(
  required_columns,
  names(monthly_data)
)

if (length(missing_columns) > 0L) {

  stop(
    paste0(
      "The following required columns are missing from monthly_data: ",
      paste(
        missing_columns,
        collapse = ", "
      )
    )
  )
}

# ================================================================
# 2. SELECT THE FIRST VALID TICKER
# ================================================================

available_tickers <- monthly_data |>
  transmute(
    ticker = as.character(ticker)
  ) |>
  filter(
    !is.na(ticker),
    ticker != ""
  ) |>
  distinct(ticker) |>
  pull(ticker)

if (length(available_tickers) == 0L) {

  stop(
    "No valid ticker was found in monthly_data."
  )
}

ticker_selected <- available_tickers[1]

# ================================================================
# 3. PREPARE ONE ADJUSTED PRICE PER MONTH
#
# If duplicate ticker-month observations exist, the final finite
# price is retained.
# ================================================================

monthly_series <- monthly_data |>
  transmute(
    ticker = as.character(ticker),

    month = as.Date(month),

    monthly_adjusted_price = suppressWarnings(
      as.numeric(
        monthly_adjusted_price
      )
    )
  ) |>
  filter(
    ticker == ticker_selected,
    !is.na(month),
    is.finite(monthly_adjusted_price)
  ) |>
  arrange(month) |>
  group_by(
    ticker,
    month
  ) |>
  summarise(
    monthly_adjusted_price = last(
      monthly_adjusted_price
    ),

    .groups = "drop"
  ) |>
  arrange(month)

n_months <- nrow(
  monthly_series
)

if (n_months < 24L) {

  stop(
    paste0(
      "At least 24 monthly observations are required for a ",
      "meaningful annual seasonal decomposition. Only ",
      n_months,
      " were found for ",
      ticker_selected,
      "."
    )
  )
}

# ================================================================
# 4. CHECK FOR DUPLICATED MONTHS
# ================================================================

duplicated_months <- monthly_series |>
  count(month) |>
  filter(
    n > 1L
  )

if (nrow(duplicated_months) > 0L) {

  stop(
    "The monthly price series contains duplicated months."
  )
}

# ================================================================
# 5. CHECK FOR MISSING CALENDAR MONTHS
#
# Classical decomposition requires regularly spaced observations.
# ================================================================

expected_month_sequence <- seq.Date(
  from = min(
    monthly_series$month
  ),

  to = max(
    monthly_series$month
  ),

  by = "month"
)

missing_calendar_months <- expected_month_sequence[
  !expected_month_sequence %in%
    monthly_series$month
]

if (length(missing_calendar_months) > 0L) {

  stop(
    paste0(
      "The monthly series contains missing calendar months: ",
      paste(
        format(
          missing_calendar_months,
          "%b %Y"
        ),
        collapse = ", "
      ),
      ". Treat the missing months before applying classical ",
      "decomposition."
    )
  )
}

# ================================================================
# 6. CREATE THE MONTHLY TIME-SERIES OBJECT
#
# frequency = 12 defines an annual seasonal cycle.
# ================================================================

first_year <- as.integer(
  format(
    min(monthly_series$month),
    "%Y"
  )
)

first_month <- as.integer(
  format(
    min(monthly_series$month),
    "%m"
  )
)

monthly_price_ts <- stats::ts(
  data = monthly_series$monthly_adjusted_price,

  start = c(
    first_year,
    first_month
  ),

  frequency = 12
)

# ================================================================
# 7. APPLY CLASSICAL ADDITIVE DECOMPOSITION
#
# Y_t = T_t + S_t + R_t
#
# stats::decompose() estimates:
#
# - trend using a centered 2 x 12 moving average;
# - seasonality by averaging detrended values by month;
# - remainder by subtraction.
# ================================================================

classical_decomposition <- stats::decompose(
  x = monthly_price_ts,

  type = "additive"
)

# ================================================================
# 8. EXTRACT THE COMPONENTS
# ================================================================

monthly_series <- monthly_series |>
  mutate(
    observed_price = as.numeric(
      classical_decomposition$x
    ),

    trend_component = as.numeric(
      classical_decomposition$trend
    ),

    seasonal_component = as.numeric(
      classical_decomposition$seasonal
    ),

    remainder_component = as.numeric(
      classical_decomposition$random
    )
  )

# ================================================================
# 9. CALCULATE THE SEASONALLY ADJUSTED SERIES
#
# For an additive decomposition:
#
# Y_t^(SA) = Y_t - S_t
#
# Therefore:
#
# Y_t^(SA) = T_t + R_t
# ================================================================

monthly_series <- monthly_series |>
  mutate(
    seasonally_adjusted_price =
      observed_price -
      seasonal_component
  )

# ================================================================
# 10. VALIDATE THE SEASONAL ADJUSTMENT
# ================================================================

if (
  any(
    !is.finite(
      monthly_series$seasonal_component
    )
  )
) {

  stop(
    "The decomposition produced invalid seasonal-component values."
  )
}

if (
  any(
    !is.finite(
      monthly_series$seasonally_adjusted_price
    )
  )
) {

  stop(
    "The seasonally adjusted series contains invalid values."
  )
}

# ================================================================
# 11. EXTRACT THE 12 UNIQUE SEASONAL INDICES
# ================================================================

monthly_series <- monthly_series |>
  mutate(
    calendar_month_number = as.integer(
      format(
        month,
        "%m"
      )
    ),

    calendar_month = factor(
      calendar_month_number,

      levels = 1:12,

      labels = month.abb,

      ordered = TRUE
    )
  )

seasonal_indices <- monthly_series |>
  group_by(
    calendar_month_number,
    calendar_month
  ) |>
  summarise(
    seasonal_effect = mean(
      seasonal_component,
      na.rm = TRUE
    ),

    .groups = "drop"
  ) |>
  arrange(
    calendar_month_number
  )

if (nrow(seasonal_indices) != 12L) {

  stop(
    paste0(
      "Expected 12 seasonal effects, but obtained ",
      nrow(seasonal_indices),
      "."
    )
  )
}

# ================================================================
# 12. VALIDATE THAT THE SEASONAL INDICES SUM TO ZERO
# ================================================================

sum_of_seasonal_effects <- sum(
  seasonal_indices$seasonal_effect
)

if (
  !isTRUE(
    all.equal(
      sum_of_seasonal_effects,
      0,
      tolerance = 1e-8
    )
  )
) {

  stop(
    paste0(
      "The additive seasonal effects do not sum to zero. ",
      "Calculated sum: ",
      format(
        sum_of_seasonal_effects,
        scientific = TRUE
      )
    )
  )
}

# ================================================================
# 13. VALIDATE THE SEASONAL-ADJUSTMENT IDENTITY
#
# Observed price =
# seasonally adjusted price + seasonal component
# ================================================================

seasonal_adjustment_error <- max(
  abs(
    monthly_series$observed_price -
      (
        monthly_series$seasonally_adjusted_price +
          monthly_series$seasonal_component
      )
  ),

  na.rm = TRUE
)

if (
  !is.finite(seasonal_adjustment_error) ||
  seasonal_adjustment_error > 1e-10
) {

  stop(
    paste0(
      "The seasonal-adjustment identity failed. ",
      "Maximum error: ",
      format(
        seasonal_adjustment_error,
        scientific = TRUE
      )
    )
  )
}

# ================================================================
# 14. VALIDATE THE FULL ADDITIVE DECOMPOSITION
#
# This validation applies only where the centered trend and
# remainder are available.
# ================================================================

complete_decomposition <- monthly_series |>
  filter(
    is.finite(trend_component),
    is.finite(remainder_component)
  )

reconstruction_error <- max(
  abs(
    complete_decomposition$observed_price -
      (
        complete_decomposition$trend_component +
          complete_decomposition$seasonal_component +
          complete_decomposition$remainder_component
      )
  ),

  na.rm = TRUE
)

if (
  !is.finite(reconstruction_error) ||
  reconstruction_error > 1e-8
) {

  stop(
    paste0(
      "The full additive decomposition identity failed. ",
      "Maximum reconstruction error: ",
      format(
        reconstruction_error,
        scientific = TRUE
      )
    )
  )
}

# ================================================================
# 15. CALCULATE THE REMOVED SEASONAL ADJUSTMENT
#
# This is the vertical distance between observed and adjusted price:
#
# Y_t - Y_t^(SA) = S_t
# ================================================================

monthly_series <- monthly_series |>
  mutate(
    adjustment_direction = case_when(
      seasonal_component > 0 ~
        "Positive seasonal effect removed",

      seasonal_component < 0 ~
        "Negative seasonal effect removed",

      TRUE ~
        "No seasonal adjustment"
    ),

    absolute_seasonal_adjustment = abs(
      seasonal_component
    )
  )

# ================================================================
# 16. IDENTIFY THE LARGEST SEASONAL ADJUSTMENTS
#
# The five largest adjustments are highlighted for teaching.
# ================================================================

n_highlight <- min(
  5L,
  nrow(monthly_series)
)

largest_adjustments <- monthly_series |>
  slice_max(
    order_by = absolute_seasonal_adjustment,

    n = n_highlight,

    with_ties = FALSE
  )

# ================================================================
# 17. SELECT ONE EXAMPLE OBSERVATION
#
# The latest occurrence of the month with the largest positive
# seasonal effect is used to illustrate the adjustment.
# ================================================================

largest_positive_month <- seasonal_indices |>
  slice_max(
    order_by = seasonal_effect,

    n = 1L,

    with_ties = FALSE
  ) |>
  pull(
    calendar_month_number
  )

example_adjustment <- monthly_series |>
  filter(
    calendar_month_number ==
      largest_positive_month
  ) |>
  slice_tail(
    n = 1L
  )

if (nrow(example_adjustment) == 0L) {

  stop(
    "The example seasonal adjustment could not be selected."
  )
}

# ================================================================
# 18. DEFINE PRICE FORMATTING
# ================================================================

median_price <- median(
  monthly_series$observed_price,

  na.rm = TRUE
)

price_accuracy <- if (
  median_price >= 100
) {
  1
} else if (
  median_price >= 10
) {
  0.1
} else {
  0.01
}

price_formatter <- label_number(
  accuracy = price_accuracy,

  big.mark = ","
)

# ================================================================
# 19. CREATE THE EXAMPLE LABEL
# ================================================================

price_range <- diff(
  range(
    c(
      monthly_series$observed_price,
      monthly_series$seasonally_adjusted_price
    ),

    na.rm = TRUE
  )
)

if (
  !is.finite(price_range) ||
  price_range <= 0
) {

  price_range <- 1
}

example_label_y <- max(
  example_adjustment$observed_price,
  example_adjustment$seasonally_adjusted_price
) +
  0.12 * price_range

example_label <- paste0(
  format(
    example_adjustment$month,
    "%b %Y"
  ),
  "\n",
  "Observed: ",
  price_formatter(
    example_adjustment$observed_price
  ),
  "\n",
  "Seasonal effect: ",
  ifelse(
    example_adjustment$seasonal_component >= 0,
    "+",
    ""
  ),
  price_formatter(
    example_adjustment$seasonal_component
  ),
  "\n",
  "Adjusted: ",
  price_formatter(
    example_adjustment$seasonally_adjusted_price
  )
)

# ================================================================
# 20. CREATE THE SEASONALLY ADJUSTED PRICE FIGURE
# ================================================================

seasonally_adjusted_plot <- ggplot(
  data = monthly_series,

  aes(
    x = month
  )
) +

  # --------------------------------------------------------------
  # Vertical segments show the seasonal adjustment:
  #
  # Observed - adjusted = seasonal component
  # --------------------------------------------------------------

  geom_segment(
    aes(
      x = month,

      xend = month,

      y = observed_price,

      yend = seasonally_adjusted_price,

      color = adjustment_direction
    ),

    linewidth = 0.65,

    alpha = 0.32
  ) +

  # --------------------------------------------------------------
  # Observed monthly adjusted price
  # --------------------------------------------------------------

  geom_line(
    aes(
      y = observed_price,

      color = "Observed monthly price"
    ),

    linewidth = 0.65,

    alpha = 0.62,

    lineend = "round"
  ) +

  # --------------------------------------------------------------
  # Seasonally adjusted price
  # --------------------------------------------------------------

  geom_line(
    aes(
      y = seasonally_adjusted_price,

      color = "Seasonally adjusted price"
    ),

    linewidth = 1.25,

    lineend = "round"
  ) +

  # --------------------------------------------------------------
  # Observed monthly points
  # --------------------------------------------------------------

  geom_point(
    aes(
      y = observed_price
    ),

    color = "#94A3B8",

    size = 1.35,

    alpha = 0.48
  ) +

  # --------------------------------------------------------------
  # Highlight selected large seasonal adjustments
  # --------------------------------------------------------------

  geom_point(
    data = largest_adjustments,

    aes(
      x = month,

      y = seasonally_adjusted_price
    ),

    inherit.aes = FALSE,

    color = "#D97706",

    fill = "white",

    shape = 21,

    size = 3.2,

    stroke = 1
  ) +

  # --------------------------------------------------------------
  # Highlight the illustrative adjustment
  # --------------------------------------------------------------

  geom_segment(
    data = example_adjustment,

    aes(
      x = month,

      xend = month,

      y = observed_price,

      yend = seasonally_adjusted_price
    ),

    inherit.aes = FALSE,

    color = "#D97706",

    linewidth = 1.5,

    arrow = grid::arrow(
      length = grid::unit(
        0.13,
        "cm"
      ),

      type = "closed"
    )
  ) +

  geom_point(
    data = example_adjustment,

    aes(
      x = month,

      y = observed_price
    ),

    inherit.aes = FALSE,

    color = "#64748B",

    fill = "white",

    shape = 21,

    size = 3.8,

    stroke = 1.1
  ) +

  geom_point(
    data = example_adjustment,

    aes(
      x = month,

      y = seasonally_adjusted_price
    ),

    inherit.aes = FALSE,

    color = "#059669",

    fill = "white",

    shape = 23,

    size = 4.2,

    stroke = 1.1
  ) +

  # --------------------------------------------------------------
  # Label the illustrative seasonal adjustment
  # --------------------------------------------------------------

  annotate(
    geom = "label",

    x = example_adjustment$month,

    y = example_label_y,

    label = example_label,

    color = "#334155",

    fill = "white",

    fontface = "bold",

    size = 3,

    lineheight = 1,

    label.size = 0.3,

    label.padding = grid::unit(
      0.22,
      "lines"
    )
  ) +

  # --------------------------------------------------------------
  # Controlled colors
  # --------------------------------------------------------------

  scale_color_manual(
    name = NULL,

    values = c(
      "Observed monthly price" =
        "#94A3B8",

      "Seasonally adjusted price" =
        "#059669",

      "Positive seasonal effect removed" =
        "#2F6FA5",

      "Negative seasonal effect removed" =
        "#DC2626",

      "No seasonal adjustment" =
        "#64748B"
    ),

    breaks = c(
      "Observed monthly price",
      "Seasonally adjusted price",
      "Positive seasonal effect removed",
      "Negative seasonal effect removed"
    )
  ) +

  # --------------------------------------------------------------
  # Date axis
  # --------------------------------------------------------------

  scale_x_date(
    date_breaks = "1 year",

    date_minor_breaks = "3 months",

    date_labels = "%Y",

    expand = expansion(
      mult = c(
        0.01,
        0.02
      )
    )
  ) +

  # --------------------------------------------------------------
  # Price axis
  # --------------------------------------------------------------

  scale_y_continuous(
    labels = price_formatter,

    expand = expansion(
      mult = c(
        0.07,
        0.19
      )
    )
  ) +

  # --------------------------------------------------------------
  # Titles
  # --------------------------------------------------------------

  labs(
    title =
      "What Changes After Removing Seasonality?",

    subtitle = paste0(
      ticker_selected,
      " monthly adjusted prices | ",
      "Seasonally adjusted series: Y(t) − S(t)"
    ),

    x = "Month",

    y = "Monthly adjusted price",

    caption = paste0(
      "The gray line is the observed monthly price. ",
      "The green line removes the estimated classical seasonal component. ",
      "Vertical segments show the adjustment applied in each month. ",
      "The seasonally adjusted series retains trend and remainder variation."
    )
  ) +

  # --------------------------------------------------------------
  # Presentation theme
  # --------------------------------------------------------------

  theme_minimal(
    base_size = 13
  ) +

  theme(
    plot.title = element_text(
      size = 20,

      face = "bold",

      color = "#172033",

      margin = margin(
        b = 4
      )
    ),

    plot.subtitle = element_text(
      size = 11.5,

      color = "#475569",

      margin = margin(
        b = 13
      )
    ),

    plot.caption = element_text(
      size = 8.5,

      color = "#64748B",

      hjust = 0,

      lineheight = 1.15,

      margin = margin(
        t = 10
      )
    ),

    axis.title = element_text(
      size = 11,

      face = "bold",

      color = "#334155"
    ),

    axis.text = element_text(
      size = 8.5,

      color = "#64748B"
    ),

    panel.grid.major = element_line(
      color = "#E5E7EB",

      linewidth = 0.4
    ),

    panel.grid.minor.x = element_line(
      color = "#F1F5F9",

      linewidth = 0.25
    ),

    panel.grid.minor.y = element_blank(),

    legend.position = "bottom",

    legend.box = "vertical",

    legend.text = element_text(
      size = 9,

      color = "#475569"
    ),

    legend.key.width = grid::unit(
      1.5,
      "cm"
    ),

    plot.margin = margin(
      t = 10,

      r = 20,

      b = 8,

      l = 10
    )
  ) +

  guides(
    color = guide_legend(
      nrow = 2,

      byrow = TRUE,

      override.aes = list(
        alpha = 1,

        linewidth = c(
          1.1,
          1.4,
          1.2,
          1.2
        )
      )
    )
  )

# ================================================================
# 21. DISPLAY THE FIGURE
# ================================================================

seasonally_adjusted_plot
```

## Time Series Decomposition


```{r}
# ================================================================
# REMAINDER COMPONENT OF MONTHLY ADJUSTED PRICES
# Classical additive decomposition
# ================================================================

library(dplyr)
library(ggplot2)
library(scales)

# ================================================================
# 1. VALIDATE THE INPUT DATASET
# ================================================================

if (!exists("monthly_data")) {

  stop(
    paste0(
      "The object 'monthly_data' does not exist. ",
      "Create the monthly adjusted-price dataset before ",
      "running this chunk."
    )
  )
}

required_columns <- c(
  "ticker",
  "month",
  "monthly_adjusted_price"
)

missing_columns <- setdiff(
  required_columns,
  names(monthly_data)
)

if (length(missing_columns) > 0L) {

  stop(
    paste0(
      "The following required columns are missing from monthly_data: ",
      paste(
        missing_columns,
        collapse = ", "
      )
    )
  )
}

# ================================================================
# 2. SELECT THE FIRST VALID TICKER
# ================================================================

available_tickers <- monthly_data |>
  transmute(
    ticker = as.character(ticker)
  ) |>
  filter(
    !is.na(ticker),
    ticker != ""
  ) |>
  distinct(ticker) |>
  pull(ticker)

if (length(available_tickers) == 0L) {

  stop(
    "No valid ticker was found in monthly_data."
  )
}

ticker_selected <- available_tickers[1]

# ================================================================
# 3. PREPARE ONE ADJUSTED PRICE PER MONTH
#
# If duplicate ticker-month observations exist, the final finite
# price is retained.
# ================================================================

monthly_series <- monthly_data |>
  transmute(
    ticker = as.character(ticker),

    month = as.Date(month),

    monthly_adjusted_price = suppressWarnings(
      as.numeric(
        monthly_adjusted_price
      )
    )
  ) |>
  filter(
    ticker == ticker_selected,
    !is.na(month),
    is.finite(monthly_adjusted_price)
  ) |>
  arrange(month) |>
  group_by(
    ticker,
    month
  ) |>
  summarise(
    monthly_adjusted_price = last(
      monthly_adjusted_price
    ),

    .groups = "drop"
  ) |>
  arrange(month)

n_months <- nrow(
  monthly_series
)

if (n_months < 24L) {

  stop(
    paste0(
      "At least 24 monthly observations are required for an ",
      "annual classical decomposition. Only ",
      n_months,
      " were found for ",
      ticker_selected,
      "."
    )
  )
}

# ================================================================
# 4. CHECK FOR DUPLICATED MONTHS
# ================================================================

duplicated_months <- monthly_series |>
  count(month) |>
  filter(
    n > 1L
  )

if (nrow(duplicated_months) > 0L) {

  stop(
    "The monthly price series contains duplicated months."
  )
}

# ================================================================
# 5. CHECK FOR MISSING CALENDAR MONTHS
#
# Classical decomposition requires a regularly spaced monthly
# series.
# ================================================================

expected_month_sequence <- seq.Date(
  from = min(
    monthly_series$month
  ),

  to = max(
    monthly_series$month
  ),

  by = "month"
)

missing_calendar_months <- expected_month_sequence[
  !expected_month_sequence %in%
    monthly_series$month
]

if (length(missing_calendar_months) > 0L) {

  stop(
    paste0(
      "The monthly series contains missing calendar months: ",
      paste(
        format(
          missing_calendar_months,
          "%b %Y"
        ),
        collapse = ", "
      ),
      ". Treat the missing months before applying classical ",
      "decomposition."
    )
  )
}

# ================================================================
# 6. CREATE THE MONTHLY TIME-SERIES OBJECT
#
# frequency = 12 defines the annual seasonal cycle.
# ================================================================

initial_year <- as.integer(
  format(
    min(monthly_series$month),
    "%Y"
  )
)

initial_month <- as.integer(
  format(
    min(monthly_series$month),
    "%m"
  )
)

monthly_price_ts <- stats::ts(
  data = monthly_series$monthly_adjusted_price,

  start = c(
    initial_year,
    initial_month
  ),

  frequency = 12
)

# ================================================================
# 7. APPLY CLASSICAL ADDITIVE DECOMPOSITION
#
# Y_t = T_t + S_t + R_t
#
# stats::decompose() estimates:
#
# - trend using a centered 2 x 12 moving average;
# - seasonality by averaging detrended prices by calendar month;
# - remainder by subtracting trend and seasonality.
# ================================================================

classical_decomposition <- stats::decompose(
  x = monthly_price_ts,

  type = "additive"
)

# ================================================================
# 8. EXTRACT THE COMPONENTS
# ================================================================

monthly_series <- monthly_series |>
  mutate(
    observed_price = as.numeric(
      classical_decomposition$x
    ),

    trend_component = as.numeric(
      classical_decomposition$trend
    ),

    seasonal_component = as.numeric(
      classical_decomposition$seasonal
    ),

    remainder_component = as.numeric(
      classical_decomposition$random
    ),

    seasonally_adjusted_price =
      observed_price -
      seasonal_component,

    detrended_price =
      observed_price -
      trend_component
  )

# ================================================================
# 9. RETAIN VALID REMAINDER OBSERVATIONS
#
# The first and final six months have no remainder because the
# centered 2 x 12 trend cannot be estimated at the sample edges.
# ================================================================

remainder_data <- monthly_series |>
  filter(
    is.finite(trend_component),
    is.finite(seasonal_component),
    is.finite(remainder_component)
  )

n_remainder_observations <- nrow(
  remainder_data
)

expected_remainder_observations <-
  n_months - 12L

if (
  n_remainder_observations !=
    expected_remainder_observations
) {

  stop(
    paste0(
      "Unexpected number of remainder observations. Expected ",
      expected_remainder_observations,
      " but obtained ",
      n_remainder_observations,
      "."
    )
  )
}

# ================================================================
# 10. VALIDATE THE ADDITIVE DECOMPOSITION
#
# Y_t = T_t + S_t + R_t
# ================================================================

reconstruction_error <- max(
  abs(
    remainder_data$observed_price -
      (
        remainder_data$trend_component +
          remainder_data$seasonal_component +
          remainder_data$remainder_component
      )
  ),

  na.rm = TRUE
)

if (
  !is.finite(reconstruction_error) ||
  reconstruction_error > 1e-8
) {

  stop(
    paste0(
      "The additive decomposition identity failed. ",
      "Maximum reconstruction error: ",
      format(
        reconstruction_error,
        scientific = TRUE
      )
    )
  )
}

# ================================================================
# 11. VALIDATE THE REMAINDER CALCULATION DIRECTLY
#
# R_t = Y_t - T_t - S_t
# ================================================================

manual_remainder <- with(
  remainder_data,

  observed_price -
    trend_component -
    seasonal_component
)

remainder_calculation_error <- max(
  abs(
    manual_remainder -
      remainder_data$remainder_component
  ),

  na.rm = TRUE
)

if (
  !is.finite(remainder_calculation_error) ||
  remainder_calculation_error > 1e-8
) {

  stop(
    paste0(
      "The manual remainder calculation does not match the ",
      "decomposition output. Maximum error: ",
      format(
        remainder_calculation_error,
        scientific = TRUE
      )
    )
  )
}

# ================================================================
# 12. CALCULATE ROBUST STANDARDIZED REMAINDERS
#
# Median and MAD are used to reduce the influence of large
# remainder observations when defining the reference scale.
# ================================================================

remainder_center <- median(
  remainder_data$remainder_component,

  na.rm = TRUE
)

remainder_scale <- mad(
  remainder_data$remainder_component,

  center = remainder_center,

  constant = 1.4826,

  na.rm = TRUE
)

# Fallback if MAD is zero or invalid

if (
  !is.finite(remainder_scale) ||
  remainder_scale <= 0
) {

  remainder_scale <- sd(
    remainder_data$remainder_component,

    na.rm = TRUE
  )
}

if (
  !is.finite(remainder_scale) ||
  remainder_scale <= 0
) {

  stop(
    "The remainder has no valid dispersion measure."
  )
}

remainder_data <- remainder_data |>
  mutate(
    robust_z_score =
      (
        remainder_component -
          remainder_center
      ) /
      remainder_scale,

    remainder_direction = case_when(
      remainder_component > 0 ~
        "Positive unexplained deviation",

      remainder_component < 0 ~
        "Negative unexplained deviation",

      TRUE ~
        "No unexplained deviation"
    ),

    remainder_class = case_when(
      abs(robust_z_score) > 3 ~
        "Highly unusual remainder",

      abs(robust_z_score) > 2 ~
        "Potentially unusual remainder",

      TRUE ~
        "Regular remainder"
    ),

    positive_remainder = if_else(
      remainder_component > 0,

      remainder_component,

      0
    ),

    negative_remainder = if_else(
      remainder_component < 0,

      remainder_component,

      0
    )
  )

# ================================================================
# 13. SEPARATE REGULAR AND UNUSUAL REMAINDERS
#
# The threshold |robust z| > 2 is used for teaching emphasis.
# It is not presented as an automatic correction rule.
# ================================================================

regular_remainders <- remainder_data |>
  filter(
    abs(robust_z_score) <= 2
  )

unusual_remainders <- remainder_data |>
  filter(
    abs(robust_z_score) > 2
  )

highly_unusual_remainders <- remainder_data |>
  filter(
    abs(robust_z_score) > 3
  )

# ================================================================
# 14. SELECT THE LARGEST POSITIVE AND NEGATIVE REMAINDERS
# ================================================================

largest_positive_remainder <- remainder_data |>
  slice_max(
    order_by = remainder_component,

    n = 1L,

    with_ties = FALSE
  )

largest_negative_remainder <- remainder_data |>
  slice_min(
    order_by = remainder_component,

    n = 1L,

    with_ties = FALSE
  )

extreme_remainders <- bind_rows(
  largest_positive_remainder,
  largest_negative_remainder
) |>
  distinct(
    month,

    .keep_all = TRUE
  )

# ================================================================
# 15. DEFINE PRICE FORMATTING
#
# Remainders retain the original price unit.
# ================================================================

median_price <- median(
  remainder_data$observed_price,

  na.rm = TRUE
)

price_accuracy <- if (
  median_price >= 100
) {
  1
} else if (
  median_price >= 10
) {
  0.1
} else {
  0.01
}

price_formatter <- label_number(
  accuracy = price_accuracy,

  big.mark = ","
)

# ================================================================
# 16. CREATE LABELS FOR THE EXTREME REMAINDERS
# ================================================================

extreme_remainders <- extreme_remainders |>
  mutate(
    extreme_label = paste0(
      format(
        month,
        "%b %Y"
      ),
      "\n",
      "Remainder: ",
      if_else(
        remainder_component >= 0,
        "+",
        ""
      ),
      price_formatter(
        remainder_component
      ),
      "\n",
      "Robust z: ",
      number(
        robust_z_score,
        accuracy = 0.01
      )
    ),

    vertical_justification = if_else(
      remainder_component >= 0,

      -0.65,

      1.55
    )
  )

# ================================================================
# 17. DEFINE A SYMMETRIC VERTICAL SCALE
#
# Symmetric limits around zero make positive and negative
# unexplained deviations directly comparable.
# ================================================================

maximum_absolute_remainder <- max(
  abs(
    remainder_data$remainder_component
  ),

  na.rm = TRUE
)

if (
  !is.finite(maximum_absolute_remainder) ||
  maximum_absolute_remainder <= 0
) {

  maximum_absolute_remainder <- 1
}

vertical_limit <-
  1.28 *
  maximum_absolute_remainder

# ================================================================
# 18. DEFINE ROBUST REFERENCE THRESHOLDS
#
# These horizontal lines correspond to:
#
# median remainder ± 2 × robust scale
# ================================================================

upper_reference <- remainder_center +
  2 * remainder_scale

lower_reference <- remainder_center -
  2 * remainder_scale

# ================================================================
# 19. CREATE THE REMAINDER FIGURE
# ================================================================

remainder_component_plot <- ggplot(
  data = remainder_data,

  aes(
    x = month,
    y = remainder_component
  )
) +

  # --------------------------------------------------------------
  # Positive remainder area
  # --------------------------------------------------------------

  geom_ribbon(
    aes(
      ymin = 0,

      ymax = positive_remainder
    ),

    fill = "#2F6FA5",

    alpha = 0.15
  ) +

  # --------------------------------------------------------------
  # Negative remainder area
  # --------------------------------------------------------------

  geom_ribbon(
    aes(
      ymin = negative_remainder,

      ymax = 0
    ),

    fill = "#DC2626",

    alpha = 0.15
  ) +

  # --------------------------------------------------------------
  # Robust reference band
  #
  # This band shows the interval:
  #
  # median remainder ± 2 × MAD-based scale
  # --------------------------------------------------------------

  geom_rect(
    aes(
      xmin = -Inf,

      xmax = Inf,

      ymin = lower_reference,

      ymax = upper_reference
    ),

    inherit.aes = FALSE,

    fill = "#64748B",

    alpha = 0.06
  ) +

  # --------------------------------------------------------------
  # Robust threshold lines
  # --------------------------------------------------------------

  geom_hline(
    yintercept = upper_reference,

    color = "#D97706",

    linewidth = 0.55,

    linetype = "dashed"
  ) +

  geom_hline(
    yintercept = lower_reference,

    color = "#D97706",

    linewidth = 0.55,

    linetype = "dashed"
  ) +

  # --------------------------------------------------------------
  # Zero reference
  #
  # R_t = 0 means that the observed price is completely explained
  # by the estimated trend and seasonal components.
  # --------------------------------------------------------------

  geom_hline(
    yintercept = 0,

    color = "#334155",

    linewidth = 0.8
  ) +

  # --------------------------------------------------------------
  # Complete remainder series
  # --------------------------------------------------------------

  geom_line(
    color = "#64748B",

    linewidth = 0.72,

    alpha = 0.82,

    lineend = "round"
  ) +

  # --------------------------------------------------------------
  # Regular remainder observations
  # --------------------------------------------------------------

  geom_point(
    data = regular_remainders,

    aes(
      x = month,
      y = remainder_component,
      color = remainder_direction
    ),

    inherit.aes = FALSE,

    size = 2,

    alpha = 0.65
  ) +

  # --------------------------------------------------------------
  # Potentially unusual remainder observations
  # --------------------------------------------------------------

  geom_point(
    data = unusual_remainders,

    aes(
      x = month,
      y = remainder_component
    ),

    inherit.aes = FALSE,

    color = "#D97706",

    fill = "white",

    shape = 21,

    size = 3.5,

    stroke = 1.05
  ) +

  # --------------------------------------------------------------
  # Highly unusual remainder observations
  # --------------------------------------------------------------

  geom_point(
    data = highly_unusual_remainders,

    aes(
      x = month,
      y = remainder_component
    ),

    inherit.aes = FALSE,

    color = "#991B1B",

    fill = "#FEE2E2",

    shape = 21,

    size = 4.2,

    stroke = 1.25
  ) +

  # --------------------------------------------------------------
  # Label the largest positive and negative remainders
  # --------------------------------------------------------------

  geom_text(
    data = extreme_remainders,

    aes(
      x = month,

      y = remainder_component,

      label = extreme_label,

      vjust = vertical_justification
    ),

    inherit.aes = FALSE,

    color = "#334155",

    size = 2.9,

    fontface = "bold",

    lineheight = 0.95,

    check_overlap = TRUE
  ) +

  # --------------------------------------------------------------
  # Point colors
  # --------------------------------------------------------------

  scale_color_manual(
    name = NULL,

    values = c(
      "Positive unexplained deviation" =
        "#2F6FA5",

      "Negative unexplained deviation" =
        "#DC2626",

      "No unexplained deviation" =
        "#059669"
    ),

    breaks = c(
      "Positive unexplained deviation",
      "Negative unexplained deviation",
      "No unexplained deviation"
    )
  ) +

  # --------------------------------------------------------------
  # Date axis
  # --------------------------------------------------------------

  scale_x_date(
    date_breaks = "1 year",

    date_minor_breaks = "3 months",

    date_labels = "%Y",

    expand = expansion(
      mult = c(
        0.01,
        0.02
      )
    )
  ) +

  # --------------------------------------------------------------
  # Symmetric remainder scale
  # --------------------------------------------------------------

  scale_y_continuous(
    limits = c(
      -vertical_limit,
      vertical_limit
    ),

    labels = price_formatter,

    breaks = pretty_breaks(
      n = 5
    ),

    expand = expansion(
      mult = c(
        0,
        0
      )
    )
  ) +

  # --------------------------------------------------------------
  # Titles and labels
  # --------------------------------------------------------------

  labs(
    title =
      "What Remains After Removing Trend and Seasonality?",

    subtitle = paste0(
      ticker_selected,
      " monthly adjusted prices | ",
      "R(t) = Y(t) − T(t) − S(t)"
    ),

    x = "Month",

    y = expression(
      hat(R)[t]
    ),

    caption = paste0(
      "Blue values represent positive unexplained deviations; ",
      "red values represent negative unexplained deviations. ",
      "Orange circles have |robust z-score| > 2, and dark-red ",
      "circles have |robust z-score| > 3. ",
      "The dashed lines show the median remainder ± 2 robust scales."
    )
  ) +

  # --------------------------------------------------------------
  # Presentation theme
  # --------------------------------------------------------------

  theme_minimal(
    base_size = 13
  ) +

  theme(
    plot.title = element_text(
      size = 20,

      face = "bold",

      color = "#172033",

      margin = margin(
        b = 4
      )
    ),

    plot.subtitle = element_text(
      size = 11.5,

      color = "#475569",

      margin = margin(
        b = 13
      )
    ),

    plot.caption = element_text(
      size = 8.5,

      color = "#64748B",

      hjust = 0,

      lineheight = 1.15,

      margin = margin(
        t = 10
      )
    ),

    axis.title = element_text(
      size = 11,

      face = "bold",

      color = "#334155"
    ),

    axis.text = element_text(
      size = 8.5,

      color = "#64748B"
    ),

    panel.grid.major = element_line(
      color = "#E5E7EB",

      linewidth = 0.4
    ),

    panel.grid.minor.x = element_line(
      color = "#F1F5F9",

      linewidth = 0.25
    ),

    panel.grid.minor.y = element_blank(),

    legend.position = "bottom",

    legend.text = element_text(
      size = 9.5,

      color = "#475569"
    ),

    legend.key.width = grid::unit(
      0.9,
      "cm"
    ),

    plot.margin = margin(
      t = 10,

      r = 20,

      b = 8,

      l = 10
    )
  ) +

  guides(
    color = guide_legend(
      override.aes = list(
        size = 3,

        alpha = 0.9
      )
    )
  )

# ================================================================
# 20. DISPLAY THE FIGURE
# ================================================================

remainder_component_plot
```


## Time Series Features

:::: {.columns}

::: {.column width="50%"}

### Trend strength

Trend strength compares the remainder with the
**trend component**:

$$
F_T =
\max\left(
0,\,
1 -
\frac{\operatorname{Var}(R_t)}
     {\operatorname{Var}(T_t + R_t)}
\right).
$$

- $F_T \approx 0$: little or no trend
- $F_T \approx 1$: strong trend

:::

::: {.column width="50%"}

### Seasonal strength

Seasonal strength compares the remainder with the
**detrended series**:

$$
F_S =
\max\left(
0,\,
1 -
\frac{\operatorname{Var}(R_t)}
     {\operatorname{Var}(S_t + R_t)}
\right)
$$

- $F_S \approx 0$: little or no seasonality
- $F_S \approx 1$: strong seasonality

:::

::::

::: {.callout-note appearance="simple" icon="false"}
Both measures range from **0 to 1**. Values closer to 1 indicate that
the corresponding systematic component explains substantially more
variation than the remainder.
:::
